---
title: "PSCC filtering input and initial sample checks"
author: "Elena Diaz"
date: "2026-06-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Setup

Load packages used for data import, table handling, REDCap access, and optional
exports.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(knitr)
library(fst)
library(writexl)
library(readr)
library(data.table)
library(tibble)
library(redcapAPI)
library(curl)
library(httr)
library(maftools)
library(DT)
library(flextable)
library(officer)
library(readxl)
library(ggplot2)
library(gridExtra)
library(R.utils)
library(tidyverse)
library(gtsummary)
```

## Custom functions in this notebook

- `save_flextable_docx()`: exports a data frame or `gtsummary` table as a
  formatted Word document.
- `normalize_name()`: standardizes column names for robust matching.
- `find_column()`: finds the first matching metadata column from a list of
  candidate names.

## Data import

All input files and external data sources used in the initial PSCC checks are
loaded here.

### Filtered variants

`list_analyzed` stores the filtered variant table produced after merging the
variant lists, removing the hypermutated sample, and removing recurrent
artefacts observed in five or more samples.

```{r import-filtered-variants}
# Load the filtered variant table used for all downstream checks.
# filtered_variants_path <- "C:/Users/ediaz/pscc_project/data/filtered_variants_outlier.fst"
filtered_variants_path <- "C:/Users/ediaz/pscc_project/data/filtered_variants.fst"

list_analyzed <- read_fst(filtered_variants_path)

# Reorder annotation columns 
list_analyzed <- list_analyzed %>%
  relocate(sift_score:gnom_ad4_1_joint_nfe_af,     .after = last_col()   )
```

### REDCap clinical table

`oncomine_followup` stores REDCap clinical information after removing personal
data fields and patients without WES samples.

```{r import-redcap}
# REDCap connection credentials are stored outside this script.
source("C:/Users/ediaz/pscc_project/tokens.R")

# Open the REDCap connection and import typed clinical records.
rcon_VSCC <- redcapConnection(
  url = api_url,
  token = token_PSCC)

oncomine_followup <- exportRecordsTyped(
  rcon = rcon_VSCC ) %>%
  select(-surnames, -sec_study_id, -age, -diag_date) %>%
  filter(!if_all(c(id_tumoral, id_premalignant, id_normal), is.na))
```

`patient_hpv` keeps the patient identifier, HPV status and tumor sample ID.

```{r import-patient-hpv}
# Keep only the identifiers needed for HPV/sample checks.
patient_hpv <- oncomine_followup %>% select(study_id, hpv_status, id_tumoral)
```

### Metadata

`metadata_pscc` stores sample-level metadata, with one row per WES sample.

```{r import-metadata}
# Load sample-level metadata.
metadata_path <- "C:/Users/ediaz/pscc_project/data/metadata-pscc.csv"
metadata_pscc <- read_delim(metadata_path, delim = ",", show_col_types = FALSE)
```

### Coverage and contamination tables

`low_coverage` and `contamination` store quality-control tables used to identify
samples that do not pass coverage or contamination thresholds.

```{r import-qc-tables}
# Load QC tables used to flag samples that fail coverage or contamination.
base_path <- "C:/Users/ediaz/pscc_project/data"

low_cov_path <- file.path(base_path, "low_coverage_samples_mutational_load.tsv")
contamination_path <- file.path(base_path, "sample_contamination_stats.tsv")

low_coverage <- read_tsv(low_cov_path, show_col_types = FALSE)
contamination <- read_tsv(contamination_path, show_col_types = FALSE)
```

### Optional exports

These exports are disabled by default.

```{r optional-exports}
# Excel format.
# write_xlsx(list_analyzed, "C:/Users/ediaz/pscc_project/downloads/list_analyzed.xlsx")

# fst format.
# write_fst(list_analyzed, "C:/Users/ediaz/pscc_project/downloads/list_analyzed.fst")
```

## General data overview

Compact overview of the main objects loaded above.

```{r data-overview}
# Print row, column, and variable-name checks for each imported object.
cat("list_analyzed:", nrow(list_analyzed), "rows (variants filtered) and", ncol(list_analyzed), "columns (variables studied)\n")
colnames(list_analyzed)

cat("oncomine_followup:", nrow(oncomine_followup), "rows and", ncol(oncomine_followup), "columns\n")
colnames(oncomine_followup)

cat("patient_hpv:", nrow(patient_hpv), "rows and", ncol(patient_hpv), "columns\n")
colnames(patient_hpv)

cat("metadata_pscc:", nrow(metadata_pscc), "rows (samples) and", ncol(metadata_pscc), "columns (variables studied)\n")
colnames(metadata_pscc)

cat("low_coverage:", nrow(low_coverage), "rows and", ncol(low_coverage), "columns\n")
colnames(low_coverage)

cat("contamination:", nrow(contamination), "rows and", ncol(contamination), "columns\n")
colnames(contamination)
```

## Saving tables and figures
example: save_flextable_docx(variant_type_counts, "variant_type_counts")
```{r}
# Export a table-like object to a formatted Word document.
save_flextable_docx <- function(data, filename) {
  
  base_path <- "C:/Users/ediaz/pscc_project/outputs/figures/"
  full_path <- paste0(base_path, filename, ".docx")
  
  # handle both dataframes and gtsummary objects
  ft <- if (inherits(data, "gtsummary")) {
    as_flex_table(data)
  } else {
    flextable(as.data.frame(data))
  }
  
  ft %>%
    theme_booktabs() %>%
    fontsize(size = 9, part = "all") %>%
    font(fontname = "Arial", part = "all") %>%
    autofit() %>%
    save_as_docx(path = full_path)
  
  message("Saved: ", full_path)
}
```

## Initial sample checks

These checks summarize the PSCC sample set and the filtered variant table used
in the downstream filtering workflow.

### HPV status

Count HPV status categories among patients with available tissue samples.

```{r hpv-status}
# Count HPV status categories in the patient/sample table.
patient_hpv %>%
  count(hpv_status)
```

### Identify key metadata columns

Detect the column names present in the metadata table.

```{r detect-columns}
# Normalize names so metadata columns can be detected despite punctuation/case differences.
normalize_name <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("_+", "_", x)
  gsub("^_|_$", "", x)
}

# Return the first metadata column matching the requested candidate names.
find_column <- function(df, candidates) {
  nm <- names(df)
  nm_clean <- normalize_name(nm)
  idx <- match(normalize_name(candidates), nm_clean)
  idx <- idx[!is.na(idx)]
  if (length(idx) == 0) return(NA_character_)
  nm[idx[1]]
}

# Detect required metadata columns before using them in summaries.
sample_col  <- find_column(metadata_pscc, c("Sample_ID", "sample_id"))
patient_col <- find_column(metadata_pscc, c("Patient", "patient"))
matched_col <- find_column(metadata_pscc, c("Matched", "matched"))
tissue_col  <- find_column(metadata_pscc, c("Tissue-type", "tissue_type", "tissue type"))

if (any(is.na(c(sample_col, patient_col, matched_col, tissue_col)))) {
  stop(
    "Could not identify one or more required metadata columns. Found columns: ",
    paste(names(metadata_pscc), collapse = ", ")
  )
}
```

### Summary of metadata samples

Summarize the metadata table at the sample level.

```{r metadata-metrics}
# Convert key metadata columns to comparable character vectors.
metadata_sample_values <- as.character(metadata_pscc[[sample_col]])
metadata_patient_values <- as.character(metadata_pscc[[patient_col]])
metadata_matched_values <- tolower(trimws(as.character(metadata_pscc[[matched_col]])))
metadata_tissue_values <- tolower(trimws(as.character(metadata_pscc[[tissue_col]])))

total_metadata_rows <- nrow(metadata_pscc)
total_unique_samples <- n_distinct(metadata_sample_values)

# Count matched and tissue-specific samples.
matched_samples_n <- sum(metadata_matched_values == "yes", na.rm = TRUE)

sample_duplicates <- tibble(sample_id = metadata_sample_values) %>%
  count(sample_id, sort = TRUE) %>%
  filter(!is.na(sample_id), n > 1)

n_tumor <- sum(metadata_tissue_values == "tumor", na.rm = TRUE)
n_tumor_matched <- sum(metadata_tissue_values == "tumor" & metadata_matched_values == "yes", na.rm = TRUE)
n_premalignant <- sum(metadata_tissue_values == "premalignant", na.rm = TRUE)
n_premalignant_matched <- sum(metadata_tissue_values == "premalignant" & metadata_matched_values == "yes", na.rm = TRUE)
n_normal <- sum(metadata_tissue_values == "normal", na.rm = TRUE)
```

```{r metadata-metrics-show}
# Assemble metadata QC metrics into a compact summary table.
metadata_summary <- tibble(
  metric = c(
    "Total rows in metadata",
    "Unique sample IDs",
    "Matched samples",
    "Tumor samples",
    "Tumor matched samples",
    "Premalignant samples",
    "Premalignant matched samples",
    "Normal samples"
  ),
  value = c(
    total_metadata_rows,
    total_unique_samples,
    matched_samples_n,
    n_tumor,
    n_tumor_matched,
    n_premalignant,
    n_premalignant_matched,
    n_normal
  )
)

metadata_summary
```

### Duplicated sample IDs in metadata

```{r metadata-duplicates}
# Report duplicate sample IDs, if any are present.
if (nrow(sample_duplicates) == 0) {
  cat("No duplicated sample IDs were found in the metadata table.\n")
} else {
  kable(sample_duplicates, caption = "Duplicated sample IDs in metadata")
}
```

### Coverage and contamination checks

Identify tumor and premalignant samples that do not pass coverage or
contamination thresholds.

```{r coverage-contamination-checks}
# Samples failing coverage threshold.
low_coverage_samples <- low_coverage %>%
  pull(sample) %>%
  unique()

# Samples failing contamination threshold.
contamination_samples <- contamination %>%
  filter(contamination > 0.05) %>%
  pull(sample) %>%
  unique()

cat(  "Low coverage samples are:",
  paste(low_coverage_samples, collapse = ", "),
  "\n" )

cat(  "Samples with contamination > 0.05 are:",
  paste(contamination_samples, collapse = ", "),
  "\n")

samples_not_passing_tum_prem <- unique(c(
  low_coverage_samples,
  contamination_samples ))

cat(  "Samples not passing coverage and/or contamination thresholds are:",
  paste(samples_not_passing_tum_prem, collapse = ", "),
  "\n" )
```

If for any reason, we want to exclude any sample_id... : 
`AS9750` was manually removed because it is hypermutated, so it is removed from
the list of passing samples without filtered variants.

```{r}
# Define manual sample exclusions.
exclude_sample_id <- c()
exclude_sample_id <- unique(c(exclude_sample_id, "AS9750"))

# Combine automatic failing samples and manually excluded samples.
samples_not_passing_all <- unique(c(
  samples_not_passing_tum_prem,
  exclude_sample_id ))

# Samples passing = non-normal samples not in failing/excluded list
metadata_passing <- metadata_pscc %>%
  filter(
    tolower(.data[[tissue_col]]) != "normal",
    !.data[[sample_col]] %in% samples_not_passing_all )

# Samples not passing = normal samples + failing/excluded tumor/premalignant samples
metadata_not_passing <- metadata_pscc %>%
  filter(
    tolower(.data[[tissue_col]]) == "normal" |
      .data[[sample_col]] %in% samples_not_passing_all )

# Tumor samples not passing
tumor_not_passing_samples <- metadata_not_passing %>%
  filter(tolower(.data[[tissue_col]]) == "tumor") %>%
  pull(.data[[sample_col]]) %>%
  unique()

# Premalignant samples not passing
premalignant_not_passing_samples <- metadata_not_passing %>%
  filter(tolower(.data[[tissue_col]]) == "premalignant") %>%
  pull(.data[[sample_col]]) %>%
  unique()
```


```{r}
# Print sample IDs assigned to the non-passing group.
cat("Tumor samples not passing are:",
  paste(tumor_not_passing_samples, collapse = ", "),
  "\n" )

cat( "Premalignant samples not passing are:",
  paste(premalignant_not_passing_samples, collapse = ", "),
  "\n" )

cat("All samples not passing are:")
metadata_not_passing
```

### Passing samples with no filtered variants

```{r passing-without-variants}
# Compare passing metadata samples against samples represented in the variant table.
passing_samples <- metadata_passing %>%
  pull(.data[[sample_col]]) %>%
  unique()

samples_with_variants <- list_analyzed %>%
  pull(sample_id) %>%
  unique()

passing_without_variants <- setdiff(
  passing_samples,
  samples_with_variants
)

if (length(passing_without_variants) > 0) {
  for (s in passing_without_variants) {
    cat(
      "Sample",
      s,
      "passes coverage and contamination thresholds but has no filtered variants.\n"
    )
  }
} else {
  cat("All passing samples have filtered variants.\n")
}
```


### Summary of filtered variants

Summarize the filtered variant table and add any passing samples with no
filtered variants.

```{r filtered-variants-summary}
# Count rows and unique samples in the filtered variant table.
total_variant_rows <- nrow(list_analyzed)

total_variant_samples <- n_distinct(list_analyzed$sample_id)

variant_tumor_samples <- list_analyzed %>%
  filter(tissue_type == "tumor") %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n)

variant_premalignant_samples <- list_analyzed %>%
  filter(tissue_type == "premalignant") %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n)

n_tumor_variants <- list_analyzed %>%
  filter(tissue_type == "tumor") %>%
  nrow()

n_premalignant_variants <- list_analyzed %>%
  filter(tissue_type == "premalignant") %>%
  nrow()

variant_summary <- tibble(
  metric = c(
    "Unique samples in filtered variants",
    "Unique tumor samples in filtered variants",
    "Unique premalignant samples in filtered variants",
    "Total variant rows",
    "Total tumor variants",
    "Total premalignant variants"
  ),
  value = c(
    total_variant_samples,
    variant_tumor_samples,
    variant_premalignant_samples,
    total_variant_rows,
    n_tumor_variants,
    n_premalignant_variants
  )
)

# Insert passing samples with no filtered variants into the summary table.
if (length(passing_without_variants) > 0) {
  rows_to_add <- tibble(
    metric = paste0(
      "Sample ",
      passing_without_variants,
      " passes but has no filtered variants"
    ),
    value = length(passing_without_variants)
  )

  variant_summary <- bind_rows(
    variant_summary %>% slice(1:3),
    rows_to_add,
    variant_summary %>% slice(4:n())
  )
}

variant_summary
```
