---
title: "TP53 variants and p53 IHC comparison"
author: "Elena Díaz"
date: "2026-03-18"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Custom functions in this notebook

No custom functions are defined in this notebook. It uses
`save_flextable_docx()` from `1_2_upload_and_initial_sample_checks.Rmd` to
export selected result tables.

## How to run

This report expects the following objects to already exist in the R environment:

- `list_analyzed`: filtered variant table 
- `oncomine_followup`: clinical REDCap table 
- `tumor_not_passing_samples`: tumor samples excluded before IHC comparison
- `patient_hpv`: patient/sample HPV table 

No extra input files are loaded directly in this notebook.

## Objective

The aim of this analysis is to assess whether tumor samples carrying **TP53
variants** in the sequencing data also show an altered **p53
immunohistochemistry (IHC)** pattern in the clinical follow-up table.

The workflow is divided into four main steps:

1.  Build an IHC table from `oncomine_followup`.
2.  Keep only tumor samples from `list_analyzed`.
3.  Classify tumor samples into **TP53-mut** and **TP53-wt** groups.
4.  Merge both groups with the IHC and HPV information.

Premalignant samples are not used here because they do not have IHC data.
Tumor matching key: `sample_id` in `list_analyzed` maps to `id_tumoral` in
`oncomine_followup`.

```{r}
# Restrict the filtered variant table to tumor samples for the IHC comparison.
list_analyzed_tumor <- list_analyzed %>%
  filter(tissue_type =="tumor")
```

#### Create IHC Table

Create a simplified IHC table using tumor sample ID and p53 IHC pattern.

```{r}
# Build a tumor-level table with simplified p53 IHC status labels.
ihc_table <- oncomine_followup %>%
  filter(!is.na(filenames_tumoral)) %>%
  select(id_tumoral, p53_ihc_status) %>%
  mutate(
    p53_ihc_status = case_when(
      p53_ihc_status == "Wt" ~ "p53-wt",
      p53_ihc_status == "Mutant" ~ "p53-ab",
      TRUE ~ NA_character_
    )
  )
# Remove tumor samples that failed filtering before comparing with variants.
ihc_table <- ihc_table %>%
  filter(!id_tumoral %in% tumor_not_passing_samples)
```

## Identify TP53-mut & TP53-wt samples

First, all rows corresponding to **TP53 variants** and **tumor tissue** are
extracted from `list_analyzed` and stored in samples_tp53mut

Second, samples considered **TP53-wt** when `sample_id` is tumor and 
appears in `metadata_passing` but absent from the TP53-mut set. 

```{r tp53-mutated-table}
# Extract tumor variants in TP53.
tp53mut <- list_analyzed_tumor %>%
  filter(gene == "TP53")

# Tumor samples with TP53 variants.
samples_tp53mut <- unique(tp53mut$sample_id)

# Get all tumor samples that passed the initial QC filters.
metadata_passing_tumor_id <- metadata_passing %>%
  filter(.data[[tissue_col]] == "tumor") %>%
  pull(Sample_ID)

# Tumor samples without TP53 variants.
samples_tp53wt <- setdiff(metadata_passing_tumor_id, samples_tp53mut)
```

```{r}
# Print sample counts and IDs for the TP53-mutated and TP53-wild-type groups.
cat("Total tumor samples:", length(metadata_passing_tumor_id), "\n")

cat("Number of TP53 variants (tumor tissue) in filtered_variants:", nrow(tp53mut), "\n")
cat("Number of unique samples with TP53 variants (tumor tissue):", n_distinct(tp53mut$sample_id), "\n")
cat("Unique sample IDs with TP53 variants (tumor tissue):\n")
print(samples_tp53mut)

cat("Total tumor samples without TP53 mutation:", length(samples_tp53wt), "\n")
cat("Unique sample IDs without TP53 variants (tumor tissue):\n")
print(unique(samples_tp53wt))
```
## Merge TP53 status info with p53 status info
```{r}
# Assign sequencing-derived TP53 status to each tumor sample in the IHC table.
ihc_genetics <- ihc_table %>%
  mutate(
    tp53_status = case_when(
      id_tumoral %in% samples_tp53wt  ~ "TP53-wt",
      id_tumoral %in% samples_tp53mut ~ "TP53-mut",
      TRUE ~ NA_character_
    )
  ) %>%
  rename(
    `Tumor Sample ID` = id_tumoral,
    `p53 IHC Status` = p53_ihc_status,
    `TP53 Status` = tp53_status
  )
ihc_genetics
```
 Discordant samples should be reviewed with additional variant annotation
sources, such as ClinVar, before drawing biological conclusions, so if filtering strategy could be enhanced.

```{r}
# Summarize concordance between p53 IHC status and TP53 variant status.
ihc_genetics %>%
  count(`p53 IHC Status`, `TP53 Status`)
```

#### Fisher Test

The contingency matrix is built directly from `ihc_genetics`: rows are IHC
groups and columns are TP53 variant groups.

```{r}
# Build the 2x2 contingency table for Fisher's exact test.
mat <- table(
  factor(ihc_genetics$`p53 IHC Status`, levels = c("p53-ab", "p53-wt")),
  factor(ihc_genetics$`TP53 Status`, levels = c("TP53-mut", "TP53-wt"))
)

rownames(mat) <- c("p53-ab", "p53-wt")
colnames(mat) <- c("TP53-mut", "TP53-wt")

mat
fisher.test(mat)

# Export the contingency table for reporting.
save_flextable_docx(mat, "tp53_ihc_contingency_table")
```

#### Add HPV Status, p53_ihc_pattern, study_id
```{r}
# Add patient, HPV, and detailed IHC pattern fields from the clinical table.
ihc_genetics_complete <- ihc_genetics %>%
  left_join(
    oncomine_followup %>%
      select(id_tumoral, study_id, hpv_status, p53_ihc_patt ) %>%
      distinct(id_tumoral, .keep_all = TRUE),
    by = c("Tumor Sample ID" = "id_tumoral")
  )

# Reorder and rename fields for a readable output table.
ihc_genetics_complete <- ihc_genetics_complete %>%
  select(study_id, `Tumor Sample ID`, `p53 IHC Status`, `TP53 Status`, hpv_status, p53_ihc_patt) %>%
  rename(
    "Study ID"                = study_id,
    "HPV Status"              = hpv_status,
    "p53 IHC Pattern (short)" = p53_ihc_patt
  )

ihc_genetics_complete

save_flextable_docx(ihc_genetics_complete, "ihc_genetics_complete")
```

```{r}
# Count HPV-associated and HPV-independent tumors in the final filtered table.
cat( "In filtered samples, HPV+:",
  sum(ihc_genetics_complete$`HPV Status` == "Associated", na.rm = TRUE),
  "\n")

cat("In filtered samples, HPV-:",
  sum(ihc_genetics_complete$`HPV Status` == "Independent", na.rm = TRUE),
  "\n")
```
