---
title: "MAF Object Creations"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Workflow Documentation

Divided into two separate files: first takes until maf object creation,
the second file records all the oncoplot generation and inspection
steps.

## Purpose

This file builds the MAF objects and oncoplots used to inspect the PSCC
filtered variant workflow. It first shows the list-level MAF summaries,
then the combined tumor/premalignant oncoplot, then the tumor-only
analysis, HPV subsets, and finally the TP53-mutational-status subsets
within HPV-independent tumors.

## Required Input Objects

Required objects already available in the R environment:

-   `list_analyzed`: final filtered variant table
-   `metadata_pscc`: sample-level metadata.
-   `oncomine_followup`: clinical table with HPV status and IHC
    variables.
-   `samples_not_passing`: samples excluded during initial sample
    checks.
-   `tp53_ihc_final`: sample-level table containing `tp53_status` from
    sequencing and `p53_ihc_status` from immunohistochemistry.

## Main Outputs

-   `maf_tumor_premalignant`: MAF object for combined tumor and
    premalignant samples from `list_analyzed`.
-   `maf_tumor`: tumor-only MAF object.
-   `maf_hpva` and `maf_hpvi`: HPV-associated and HPV-independent tumor
    MAF objects.
-   `maf_hpvi_tp53wt` and `maf_hpvi_tp53mut`: HPV-independent tumor MAF
    objects split by sequencing-derived TP53 mutational status.

## Important Assumptions

-   `tp53_status` means sequencing-derived TP53 status.
-   `p53_ihc_status` means p53 immunohistochemistry pattern.
-   Normal samples are excluded from this oncoplot workflow.
-   Samples listed in `samples_not_passing` are excluded from the
    clinical annotation table.
-   Samples with clinical annotation but no filtered variants are
    retained in oncoplots using silent placeholder rows.

## Custom functions in this notebook

- `prepare_maf_table()`: converts the filtered variant table into the column
  structure expected by `maftools`.
- `add_dummy_samples()`: adds silent placeholder rows for clinically retained
  samples with no filtered variants.
- `top_genes()`: returns the top mutated genes from a MAF object.
- `build_maf_from_clinical()`: builds a MAF object for a selected clinical
  subset.
- `get_ordered_genes()`: orders genes for the grouped tumor oncoplot, giving
  priority to HPV-independent TP53-mutated tumors.

# 1. Helper Functions and Shared Settings

```{r helpers}
# Map Sequence Ontology annotations to maftools variant classes.
so_to_maf <- c(
  "missense_variant" = "Missense_Mutation",
  "missense_variant&splice_region_variant" = "Missense_Mutation",
  "frameshift_variant" = "Frame_Shift_Ins",
  "stop_gained" = "Nonsense_Mutation",
  "stop_gained&splice_region_variant" = "Nonsense_Mutation",
  "stop_lost" = "Nonstop_Mutation",
  "splice_donor_variant&intron_variant" = "Splice_Site",
  "splice_donor_variant&splice_region_variant&intron_variant" = "Splice_Site",
  "splice_acceptor_variant&intron_variant" = "Splice_Site",
  "splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant" = "Splice_Site",
  "splice_donor_variant&missense_variant&splice_region_variant&intron_variant" = "Splice_Site",
  "splice_acceptor_variant&missense_variant&splice_region_variant&intron_variant" = "Splice_Site",
  "frameshift_variant&synonymous_variant" = "Frame_Shift_Del",
  "frameshift_variant&missense_variant" = "Frame_Shift_Ins",
  "start_lost" = "Nonsense_Mutation"
)

# Convert filtered variants into a MAF-compatible data.table.
prepare_maf_table <- function(variant_dt) {
  maf_table <- copy(as.data.table(variant_dt))

  if ("tissue-type" %in% names(maf_table)) {
    setnames(maf_table, "tissue-type", "tissue_type")
  }

  maf_table[, Tumor_Sample_Barcode := as.character(sample_id)]
  maf_table[, Hugo_Symbol := as.character(gene)]
  maf_table[, Chromosome := as.character(number_chrom)]
  maf_table[, Start_Position := as.integer(pos)]
  maf_table[, End_Position := as.integer(pos)]
  maf_table[, Reference_Allele := as.character(ref)]
  maf_table[, Tumor_Seq_Allele2 := as.character(alt)]
  maf_table[, Variant_Classification := so_to_maf[annotation]]

  # Reclassify frameshift events as insertions or deletions based on allele length.
  maf_table[
    !is.na(annotation) & grepl("frameshift", annotation, ignore.case = TRUE),
    Variant_Classification := fifelse(
      nchar(Tumor_Seq_Allele2) > nchar(Reference_Allele),
      "Frame_Shift_Ins",
      "Frame_Shift_Del"
    )
  ]

  # Report annotation terms that do not map directly to maftools classes.
  unmapped <- unique(maf_table[is.na(Variant_Classification), annotation])
  if (length(unmapped) > 0) {
    message(
      "Unmapped annotation terms, kept as NA and handled by maftools: ",
      paste(unmapped, collapse = ", ")
    )
  }

  # Derive the maftools variant type from reference and alternate allele lengths.
  maf_table[, ref_len := nchar(Reference_Allele)]
  maf_table[, alt_len := nchar(Tumor_Seq_Allele2)]
  maf_table[, Variant_Type := fcase(
    Reference_Allele == "-" | alt_len > ref_len, "INS",
    Tumor_Seq_Allele2 == "-" | alt_len < ref_len, "DEL",
    ref_len == 1L, "SNP",
    ref_len == 2L, "DNP",
    ref_len == 3L, "TNP",
    default = "ONP"
  )]
  maf_table[, c("ref_len", "alt_len") := NULL]

  maf_table
}

# Add silent rows so samples without retained variants remain visible in plots.
add_dummy_samples <- function(maf_table, clinical_dt) {
  samples_no_variants <- setdiff(
    clinical_dt$Tumor_Sample_Barcode,
    maf_table$Tumor_Sample_Barcode
  )

  if (length(samples_no_variants) == 0) {
    return(maf_table)
  }

  dummy_rows <- data.table(
    Tumor_Sample_Barcode = as.character(samples_no_variants),
    Hugo_Symbol = "Unknown",
    Chromosome = "1",
    Start_Position = 1L,
    End_Position = 1L,
    Reference_Allele = "A",
    Tumor_Seq_Allele2 = "A",
    Variant_Classification = "Silent",
    Variant_Type = "SNP"
  )

  rbindlist(list(maf_table, dummy_rows), fill = TRUE)
}

# Return the most frequently mutated genes from a MAF object.
top_genes <- function(maf_object, n = 10) {
  gene_summary <- getGeneSummary(maf_object)
  gene_summary[seq_len(min(n, nrow(gene_summary))), Hugo_Symbol]
}
```

# 2. Prepare Variant Tables

```{r prepare-variant-tables}
# Create MAF-ready and data.table versions of the filtered variants.
  maf_table_all <- prepare_maf_table(list_analyzed)

  filtered_variants <- copy(as.data.table(list_analyzed))
```

# 3. Prepare Clinical Annotation Table
Prepare all data in metadata filtered

```{r clinical-annotation}
# Merge sample metadata with clinical follow-up variables.
metadata_filtered <- metadata_pscc %>%
  left_join(oncomine_followup, by = c("Patient" = "study_id")) %>%
  rename_with(~ gsub("-", "_", tolower(.x)))

samples_not_passing_ids <- as.character(metadata_not_passing$Sample_ID)

# Keep tumor and premalignant samples that passed the initial QC filters.
metadata_filtered <- metadata_filtered %>%
  filter(
    !sample_id %in% samples_not_passing_ids,
    tissue_type %in% c("tumor", "premalignant")  )

# Add TP53 sequencing status and p53 IHC status from the IHC comparison table.
metadata_filtered <- metadata_filtered %>%
  select(-p53_ihc_status) %>%
  left_join(
    ihc_genetics %>%
      select(
        `Tumor Sample ID`,
        `TP53 Status`,
        `p53 IHC Status`
      ),
    by = c("sample_id" = "Tumor Sample ID")
  )

metadata_filtered <- metadata_filtered %>%
  rename(
    tp53_status = `TP53 Status`, 
    p53_ihc_status = `p53 IHC Status`
  )

colnames(metadata_filtered)
```

```{r clinical-recoding}
# Select clinical variables used by maftools annotations.
clinical_for_maf_all <- as.data.table(metadata_filtered)[,.(
    Tumor_Sample_Barcode = as.character(sample_id),
    patient, hpv_status, tissue_type, stage, p53_ihc_status, 
    tp53_status, cyclind1_ihc_perc, matched, age_calc)]

clinical_for_maf_all <- unique(clinical_for_maf_all)

# Add minimal clinical rows for variant samples missing from the merged metadata.
missing_clinical_samples <- setdiff(
  filtered_variants$sample_id,
  clinical_for_maf_all$Tumor_Sample_Barcode)

if (length(missing_clinical_samples) > 0) {
  clinical_for_maf_all <- rbindlist(
    list(
      clinical_for_maf_all,
      unique(filtered_variants[
        sample_id %in% missing_clinical_samples,
        .(
          Tumor_Sample_Barcode = sample_id,
          patient,
          tissue_type,
          matched
        )
      ])
    ),
    fill = TRUE
  )
}

# Confirm how many clinical samples remain for MAF construction.
cat(  "Clinical samples retained:",
  uniqueN(clinical_for_maf_all$Tumor_Sample_Barcode),
  "\n" )
```

# 4. Recode clinical variables
Standardize follow-up variables before downstream clinical annotation
uses.

```{r clinical-recoding}
# Ensure numeric fields are typed before plotting or binning.
clinical_for_maf_all[, age_calc := as.numeric(age_calc)]


# Identify samples with sequencing-detected TP53 variants.
tp53mut_samples <- unique(
  filtered_variants[gene == "TP53", sample_id]
)

# Standardize HPV status labels.
clinical_for_maf_all[, hpv_status := fcase(
  hpv_status %in% c("Associated", "associated"), "Associated",
  hpv_status %in% c("Independent", "independent"), "Independent",
  hpv_status %in% c("Unknown", "unknown", "NA"), "Unknown",
  default = NA_character_
)]

# Standardize TP53 status labels.
clinical_for_maf_all[, tp53_status := fcase(
  tp53_status %in% c("TP53-mut", "TP53mut", "TP53-mutated"), "TP53-mutated",
  tp53_status %in% c("TP53-wt", "TP53wt", "TP53-wild-type"), "TP53-wild-type",
  default = NA_character_
)]

# Fill missing TP53 status using the filtered variant table.
clinical_for_maf_all[
  is.na(tp53_status) | tp53_status == "",
  tp53_status := fifelse(
    Tumor_Sample_Barcode %in% tp53mut_samples,
    "TP53-mutated",
    "TP53-wild-type"
  )
]

# Standardize p53 IHC labels.
clinical_for_maf_all[, p53_ihc_status := fcase(
  p53_ihc_status %in% c("p53-ab", "p53-abnormal", "p53 abnormal"), "p53-aberrant",
  p53_ihc_status %in% c("p53-wt", "p53-wild-type", "p53 wild type"), "p53-wild-type",
  default = NA_character_
)]

# Recode remaining clinical variables for consistent oncoplot annotations.
clinical_for_maf_all[, stage := as.integer(stage)]
clinical_for_maf_all[, cyclind1_ihc_perc := as.numeric(cyclind1_ihc_perc)]
clinical_for_maf_all[, matched := fcase(
  tolower(trimws(as.character(matched))) == "yes", "Yes",
  tolower(trimws(as.character(matched))) == "no", "No",
  default = NA_character_
)]
clinical_for_maf_all[, tissue_type := trimws(tolower(as.character(tissue_type)))]

# Convert cyclin D1 IHC percentage into grouped annotation bins.
clinical_for_maf_all[, cyclind1_ihc_perc_grp := cut(
  cyclind1_ihc_perc,
  breaks = c(-Inf, 0, 5, 10, 25, 50, 75, 100),
  labels = c("0", "1-5", "6-10", "11-25", "26-50", "51-75", "76-100"),
  right = TRUE
)]

# Define the combined HPV/TP53 grouping used for tumor oncoplot ordering.
clinical_for_maf_all[, HPV_TP53_mutational_group := fcase(
  hpv_status == "Associated",
  "HPV-A",
  hpv_status == "Independent" & tp53_status == "TP53-mutated",
  "HPV-I/TP53mut",
  hpv_status == "Independent" & tp53_status == "TP53-wild-type",
  "HPV-I/TP53wt",
  default = NA_character_
)]

# Set annotation factor levels to control display order in plots.
clinical_for_maf_all[, hpv_status := factor(hpv_status, levels = c("Associated", "Independent", "Unknown"))]
clinical_for_maf_all[, tissue_type := factor(tissue_type, levels = c("tumor", "premalignant"))]
clinical_for_maf_all[, stage := factor(stage, levels = c(1, 2, 3, 4), labels = c("I", "II", "III", "IV"))]
clinical_for_maf_all[, tp53_status := factor(tp53_status, levels = c("TP53-wild-type", "TP53-mutated"))]
clinical_for_maf_all[, p53_ihc_status := factor(p53_ihc_status, levels = c("p53-wild-type", "p53-aberrant"))]
clinical_for_maf_all[, HPV_TP53_mutational_group := factor(
  HPV_TP53_mutational_group,
  levels = c("HPV-A", "HPV-I/TP53mut", "HPV-I/TP53wt")
)]
clinical_for_maf_all[, matched := factor(matched, levels = c("No", "Yes"))]

# Keep a tumor-only clinical table for tumor-specific MAF objects.
clinical_for_maf <- copy(clinical_for_maf_all[tissue_type == "tumor"])

```

#5. Build MAF Objects maf_tumor object and maf_tumor_premalignant

```{r build-maf-objects}
# Keep only samples present in the clinical annotation table.
# Then add silent placeholder rows for samples with no retained variants.
maf_table_all <- maf_table_all[
  Tumor_Sample_Barcode %in% clinical_for_maf_all$Tumor_Sample_Barcode
]
maf_table_all <- add_dummy_samples(maf_table_all, clinical_for_maf_all)

# Tumor-only MAF table, including dummy rows already added above.
maf_table_tumor <- maf_table_all[
  Tumor_Sample_Barcode %in% clinical_for_maf$Tumor_Sample_Barcode
]
```

```{r}
# Build a MAF object from a clinical subset and matching mutation rows.
build_maf_from_clinical <- function(clinical_dt, maf_dt) {
  sample_ids <- as.character(clinical_dt$Tumor_Sample_Barcode)

  read.maf(
    maf = maf_dt[Tumor_Sample_Barcode %in% sample_ids],
    clinicalData = as.data.frame(clinical_dt)
  )
}

# Build the combined tumor/premalignant and tumor-only MAF objects.
maf_tumor_premalignant <- build_maf_from_clinical(
  clinical_for_maf_all,
  maf_table_all
)

maf_tumor <- build_maf_from_clinical(
  clinical_for_maf,
  maf_table_tumor
)

# Build HPV-associated and HPV-independent tumor MAF subsets.
maf_hpva <- build_maf_from_clinical(
  clinical_for_maf[as.character(hpv_status) == "Associated"],
  maf_table_tumor
)

maf_hpvi <- build_maf_from_clinical(
  clinical_for_maf[as.character(hpv_status) == "Independent"],
  maf_table_tumor
)

# Build HPV-independent tumor subsets by sequencing-derived TP53 status.
maf_hpvi_tp53wt <- build_maf_from_clinical(
  clinical_for_maf[
    as.character(hpv_status) == "Independent" &
      as.character(tp53_status) == "TP53-wild-type"
  ],
  maf_table_tumor
)

maf_hpvi_tp53mut <- build_maf_from_clinical(
  clinical_for_maf[
    as.character(hpv_status) == "Independent" &
      as.character(tp53_status) == "TP53-mutated"
  ],
  maf_table_tumor
)
```

```{r}
# Store all generated MAF objects in a named list for summaries.
maf_objects <- list(
  tumor_premalignant = maf_tumor_premalignant,
  tumor = maf_tumor,
  hpva = maf_hpva,
  hpvi = maf_hpvi,
  hpvi_tp53wt = maf_hpvi_tp53wt,
  hpvi_tp53mut = maf_hpvi_tp53mut
)
```

```{r}
# Summarize the clinical sample count represented in each MAF object.
maf_build_summary <- data.table(
  maf_object = names(maf_objects),
  clinical_samples = c(
    uniqueN(clinical_for_maf_all$Tumor_Sample_Barcode),
    uniqueN(clinical_for_maf$Tumor_Sample_Barcode),
    uniqueN(clinical_for_maf[as.character(hpv_status) == "Associated", Tumor_Sample_Barcode]),
    uniqueN(clinical_for_maf[as.character(hpv_status) == "Independent", Tumor_Sample_Barcode]),
    uniqueN(clinical_for_maf[
      as.character(hpv_status) == "Independent" &
        as.character(tp53_status) == "TP53-wild-type",
      Tumor_Sample_Barcode
    ]),
    uniqueN(clinical_for_maf[
      as.character(hpv_status) == "Independent" &
        as.character(tp53_status) == "TP53-mutated",
      Tumor_Sample_Barcode
    ])
  ))

maf_build_summary
```

```{r}
# Print maftools sample summaries for each MAF object.
getSampleSummary(maf_tumor_premalignant)
getSampleSummary(maf_tumor)
getSampleSummary(maf_hpva)
getSampleSummary(maf_hpvi)
getSampleSummary(maf_hpvi_tp53wt)
getSampleSummary(maf_hpvi_tp53mut)
```

# 6. For the general oncoplot
```{r}
# Define tumor sample groups used in the grouped oncoplot.
samples_hpva <- clinical_for_maf[
  as.character(hpv_status) == "Associated",
  Tumor_Sample_Barcode
]

samples_hpvi <- clinical_for_maf[
  as.character(hpv_status) == "Independent",
  Tumor_Sample_Barcode
]

samples_hpvi_tp53wt <- clinical_for_maf[
  Tumor_Sample_Barcode %in% samples_hpvi &
    as.character(tp53_status) == "TP53-wild-type",
  Tumor_Sample_Barcode
]
samples_hpvi_tp53mut <- clinical_for_maf[
  Tumor_Sample_Barcode %in% samples_hpvi &
    as.character(tp53_status) == "TP53-mutated",
  Tumor_Sample_Barcode
]
```

```{r}
# Sample order --> HPVA, HPVI-TP53mut, HPVI-TP53wt
ordered_samples <- unique(c(
  samples_hpva,
  samples_hpvi_tp53mut,
  samples_hpvi_tp53wt))

group_order <- c(
  "HPV-A",
  "HPV-I/TP53mut",
  "HPV-I/TP53wt")

# Build clinical data in the same order as the sample display order.
clinical_three_groups <- as.data.frame(
  clinical_for_maf[Tumor_Sample_Barcode %in% ordered_samples])

clinical_three_groups <- clinical_three_groups[match(
  ordered_samples,
  clinical_three_groups$Tumor_Sample_Barcode), ]

clinical_three_groups$HPV_TP53_mutational_group <- factor(
  as.character(clinical_three_groups$HPV_TP53_mutational_group),
  levels = group_order
)

# Build a grouped tumor MAF object using the ordered clinical table.
maf_three_groups <- read.maf(
  maf = maf_table_tumor[Tumor_Sample_Barcode %in% ordered_samples],
  clinicalData = clinical_three_groups
)
```

```{r}
# Gene order --> set by HPVI-TP53mut
top_genes_global <- getGeneSummary(maf_three_groups)$Hugo_Symbol

hpvi_tp53mut_gene_counts <- sort(table(unique(
  as.data.frame(maf_hpvi_tp53mut@data[, c("Hugo_Symbol", "Tumor_Sample_Barcode")])
)$Hugo_Symbol), decreasing = TRUE)

# Return the top genes with HPV-I/TP53-mutated genes shown first when present.
get_ordered_genes <- function(n_top_genes) {
  
  top_genes <- top_genes_global[seq_len(min(n_top_genes, length(top_genes_global)))]
  
  hpvi_tp53mut_gene_counts <- sort(table(unique(
    as.data.frame(maf_hpvi_tp53mut@data[, c("Hugo_Symbol", "Tumor_Sample_Barcode")])
  )$Hugo_Symbol), decreasing = TRUE)
  
  ordered_genes <- c(
    intersect(names(hpvi_tp53mut_gene_counts), top_genes),
    setdiff(top_genes, names(hpvi_tp53mut_gene_counts))
  )
  
  ordered_genes[!is.na(ordered_genes)]
}
```

# 7. Color annotations

```{r annotation-colors}
# Define fixed colors for all clinical annotation tracks used in oncoplots.
oncoplot_ann_colors <- list(
  HPV_TP53_mutational_group = c(
    "HPV-A" = "#377EB8",
    "HPV-I/TP53mut" = "#E69F00",
    "HPV-I/TP53wt" = "#009E73"
  ),
  tissue_type = c(
    "tumor" = "#7B3294",
    "premalignant" = "#A6DBA0"
  ),
  hpv_status = c(
    "Associated" = "#2C7BB6",
    "Independent" = "#FDAE61",
    "Unknown" = "#BDBDBD"
  ),
  tp53_status = c(
    "TP53-mutated" = "#CC79A7",
    "TP53-wild-type" = "#56B4E9"
  ),
  p53_ihc_status = c(
    "p53-aberrant" = "#D55E00",
    "p53-wild-type" = "#0072B2"
  ),
  matched = c(
    "Yes" = "#4D9221",
    "No" = "#762A83"
  ),
  cyclind1_ihc_perc_grp = c(
    "0" = "grey95",
    "1-5" = "grey85",
    "6-10" = "grey75",
    "11-25" = "grey65",
    "26-50" = "grey55",
    "51-75" = "grey45",
    "76-100" = "grey35"
  )
)
```

