---
title: "Cyclin D1 IHC Boxplot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Custom functions used in this notebook

No custom functions are defined in this notebook. It uses `clinical_for_maf`,
which is created upstream, to summarize and compare Cyclin D1 IHC percentages
across HPV/TP53 groups.

# Workflow overview

This notebook summarizes Cyclin D1 IHC percentages by
`HPV_TP53_mutational_group`, plots the group distributions, and runs global and
pairwise non-parametric tests.

#clinical_for_maf data
We only use HPV_TP53_mutational group and cyclind1_ihc_perc

```{r cyclind1-summary}
# Summarize Cyclin D1 IHC percentages within each HPV/TP53 group.
clinical_for_maf %>%
  group_by(HPV_TP53_mutational_group) %>%
  summarise(
    n = n(),
    median = median(cyclind1_ihc_perc),
    q1 = quantile(cyclind1_ihc_perc, 0.25),
    q3 = quantile(cyclind1_ihc_perc, 0.75),
    min = min(cyclind1_ihc_perc),
    max = max(cyclind1_ihc_perc),
    .groups = "drop"
  )
```

```{r cyclind1-boxplot, fig.width=6, fig.height=4, dpi=300}
# Plot Cyclin D1 IHC percentages by HPV/TP53 group.
ccd1_ihc_by_hpvtp53groups <- ggplot(clinical_for_maf, aes(x = HPV_TP53_mutational_group, y = cyclind1_ihc_perc, fill = HPV_TP53_mutational_group)) +
  geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.75) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.85, shape = 21, color = "grey20") +
  scale_fill_manual(
    values = c(
      "HPV-I/TP53wt" = "#009E73",
      "HPV-I/TP53mut" = "#E69F00",
      "HPV-A" = "#377EB8"
    )
  ) +
  scale_y_continuous(breaks = seq(0, 100, 25), limits = c(0, 100)) +
  labs(
    x = NULL,
    y = "Cyclin D1 IHC (%)",
    title = "Cyclin D1 IHC by HPV/TP53 group"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

ccd1_ihc_by_hpvtp53groups
```

```{r cyclind1-kruskal}
# Test for global Cyclin D1 IHC differences across HPV/TP53 groups.
ccdn1_kruskal <- kruskal.test(cyclind1_ihc_perc ~ HPV_TP53_mutational_group, data = clinical_for_maf)
ccdn1_kruskal
```

```{r cyclind1-pairwise-wilcox}
# Run pairwise Wilcoxon comparisons with BH-adjusted p-values.
ccdn1_wilcox <- pairwise.wilcox.test(
  x = clinical_for_maf$cyclind1_ihc_perc,
  g = clinical_for_maf$HPV_TP53_mutational_group,
  p.adjust.method = "BH",
  exact = FALSE
)
ccdn1_wilcox
```
