---
title: "cohort_table"
author: "Elena Díaz"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Custom functions used in this notebook

No custom functions are defined in this notebook. It uses `save_flextable_docx()`,
which is created upstream, to export the final cohort table.

# Workflow overview

This script is explicity created to prepare the cohort table 
All information for the rows has been taken from RedCap. Stored in oncomine_followup
The three groups of the columns are taken from HPV_TP53_mutational_group from clinical_for_maf
Doubt: all cohort or only the patients from whom we have data? We do not have genetic
TP53 data from patients who are missing coverage/contamination
- All cohort: p53 status by IHC
- Only passing patients: tp53 genetic status can be used




```{r}
# Join REDCap follow-up variables with HPV/TP53 group annotations.
cohort_table_all <- oncomine_followup %>%
  select(study_id, age_calc, stage, hist_type, radiotherapy, chemotherapy,status_last_followup, recurrence, p53_ihc_status ) %>%
rename(patient = study_id) %>%
  left_join(clinical_for_maf %>% select(patient, HPV_TP53_mutational_group), by = "patient")
  
  cohort_table_all
  
```


```{r}
# Recode clinical variables and generate the grouped summary table.
cohort <- cohort_table_all %>%
  rename(age = age_calc) %>%
  select(!c(patient)) %>%
  filter(
    stage != "0",
    radiotherapy != "Unknown"
  ) %>%
  mutate(
    stage = case_when(
      stage %in% c("I", "II") ~ "Initial (I-II)",
      stage %in% c("III", "IV") ~ "Advanced (III-IV)",
      TRUE ~ stage
    ),
    hist_type = case_when(
      hist_type %in% c("Basaloid", "Warty", "Nonkeratinizing") ~ 
        "Basaloid/Warty/Nonkeratinizing",
      hist_type %in% c("Usual", "Keratinizing", "Lymphoepithelioma-like", "Mixed") ~ 
        "Usual/Keratinizing",
      TRUE ~ hist_type
    )
  ) %>%
  tbl_summary(
              by = HPV_TP53_mutational_group, 
              label = list( age ~ "Age", 
                            stage ~ "Stage", 
                            hist_type ~ "Histological Type"),
              statistic = list(
                            age ~ "{mean} ({min}–{max})",
                            all_continuous() ~ "{mean} ({sd})", 
                            all_categorical() ~ "{n} / {N} ({p}%)"
                            ), 
                             digits = list(
                              age ~ 1,
                              all_continuous() ~ 1,
                               all_categorical() ~ c(0, 0, 1)
    ),
    missing = "no" ) %>%
  add_p(test = list(all_continuous() ~ "kruskal.test", all_categorical() ~ "fisher.test")) %>%
  bold_labels() %>%  add_overall() 

cohort
```


```{r}
# Export the cohort summary table to Word.
save_flextable_docx(cohort, "cohort")
```

# Save the current R workspace for downstream notebooks.
save.image(
  file = "C:/Users/ediaz/pscc_project/2.filtering/my_scripts/maf_workspace.RData"
)


