---
title: "Pathway-Level Mutation Burden"
author: "Elena Diaz"
date: "2026-05-11"
output: html_document
---

# Custom functions defined in this notebook

- `make_pairwise_pathway_boxplots()`: builds pathway-level boxplots and
  Wilcoxon summary tables for a specified two-group comparison.

# Custom functions and objects used from upstream notebooks

- `maf_tumor`: tumor-only MAF object used for pathway oncoplots and mutation
  counts.
- `clinical_for_maf`: clinical annotation table used to assign samples to
  HPV/TP53 groups.
- `ordered_samples`: sample order created upstream for grouped oncoplots.
- `oncoplot_ann_colors`: annotation color list used by the pathway oncoplots.
- `group_order`: fixed HPV/TP53 group order used in pathway boxplots.

# Workflow Documentation

## Purpose

This notebook assigns recurrently mutated tumor genes to functional pathways.

We have threee pathways lists:
1&2. The easiest with sigpw, smgbp which are a oncogenic signalling pathwat and a catalog
of known genes classified by their biological processes.
3. Using the list provided by Robert

The representations will be sample ordered by HPB_TP53_groups, to check if there is any conclusion
- `HPVA`
- `HPVI TP53-mutated`
- `HPVI TP53-wild-type`
Then we go for grouping by nº of mutations per pathway

It uses `maftools::oncoplot()` to display pathway annotations on the tumor MAF
and then creates one boxplot per pathway with Kruskal-Wallis tests across
`HPV_TP53_mutational_group`.

## Required Previous Steps

Run these files first in the same R session:

1. `1upload_to_R.Rmd initial_sample_checks.Rmd`
3. `3tp53_ihc_variants.Rmd`
4. `4merged_oncoplot_workflow.Rmd`

## Required Objects in the R Environment

This notebook expects the following objects from the previous workflow:

- `maf_tumor`: tumor-only MAF object created in `4merged_oncoplot_workflow.Rmd`.
- `clinical_for_maf`: tumor-only clinical annotation table.
- `oncoplot_ann_colors`: optional color list for the clinical annotations. If it
  is missing, local fallback colors are used.

## Required File on Disk

The pathway annotation file must be available here:

`C:/Users/ediaz/pscc_project/server/pathways/gene_annotations_recurrence_3.xlsx`

It must contain the columns `Gene` and `Pathway2`.


# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Recurrent Genes Used for Pathway Counts

The analysis is restricted to recurrent tumor genes mutated in at least 3
samples. This matches the gene set shown in the pathway oncoplot and the genes
appearing in Robert_pathway dataset.

```{r recurrent-genes}
# Identify recurrent tumor genes mutated in at least three samples.
gene_summary <- getGeneSummary(maf_tumor) %>%
  as_tibble()

recurrent_genes <- gene_summary %>%
  filter(MutatedSamples >= 3) %>%
  arrange(desc(MutatedSamples)) %>%
  pull(Hugo_Symbol)

length(recurrent_genes)
recurrent_genes
```

# 2. Load Pathway Annotation

```{r load-pathways}
# Load Robert pathway annotations for the recurrent gene set.
pathway_file <- "C:/Users/ediaz/pscc_project/server/pathways/gene_annotations_recurrence_3.xlsx"
Robert_pathways <- read_excel(pathway_file) %>%
  select(
    gene,
    general_category)
#as this list is taken exactly for my genes, this table provides info of my gene cohort
Robert_pathways %>%
  count(general_category, name = "n_genes") %>%
  arrange(desc(n_genes))
```

# Grouping by pathways
The simplest plot using each of the pathway databases from maftools:
- oncogenic signalling pathways: A curated list of 10 signalling pathways
- Catalog of 10 known driver genes classified by their biological processes
- For the Robert_Database we have 11 gene categories
For the two first plt0s we set top 84 genes that is the number for recurrence => 3


# Adding order by HPV_TP53_mutational_group
Now we plot the same but grouping the samples (x axis) by annotation of HPV_TP53_mutational_group. 

Now we plot the three oncoplots by ordering samples by HPVA, HPVi-TP53mut,  HPVi-TP53wt
```{r, fig.width=16, fig.height=12, dpi=300, fig.retina=2}
# Plot pathway annotations using the maftools sigpw database.
genes_pathways_sigpw <- oncoplot(maf= maf_tumor, pathways = "sigpw", gene_mar =8, fontSize=0.6, topPathways= 10, top = 84, 
        sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = FALSE,
        showTumorSampleBarcodes = TRUE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        annotationFontSize = 1.1,
        legendFontSize = 0.9,
        titleFontSize = 1.3,
        SampleNamefontSize = 0.75)

# Plot pathway annotations using the maftools smgbp database.
genes_pathways_smgbp <- oncoplot(maf= maf_tumor, pathways = "smgbp", gene_mar =8, fontSize=0.4, topPathways= 10, top = 84,
        sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = FALSE,
        showTumorSampleBarcodes = TRUE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        annotationFontSize = 0.5,
        legendFontSize = 0.9,
        titleFontSize = 1.3,
        SampleNamefontSize = 0.5)

# Plot pathway annotations using Robert pathway categories.
genes_pathways_rob <- oncoplot(maf= maf_tumor, pathways = Robert_pathways, gene_mar =8, fontSize=0.5,topPathways= 11, 
        top = 84, sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = FALSE,
        showTumorSampleBarcodes = TRUE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        annotationFontSize = 1.1,
        legendFontSize = 0.9,
        titleFontSize = 1.3,
        SampleNamefontSize = 0.75)

genes_pathways_sigpw
#genes_pathways_smgbp
#genes_pathways_rob
```

# Grouping by pathways &  number of mutations per pathway
We first plot the same plot as first and add collapsePathway.

Adding to this representation the differenciation between groups HPV-TP53.
This is chunck 8 adding , collapsePathway = TRUE, sortByAnnotation = TRUE,
```{r}
# Collapse sigpw genes into pathway-level mutation rows.
pathways_sigpw <- oncoplot(maf= maf_tumor, pathways = "sigpw", gene_mar =8, fontSize=0.6, topPathways= 10, top = 84, 
        sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = TRUE,
        showTumorSampleBarcodes = FALSE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        barcode_mar = 5,
        anno_height = 0.5, 
        collapsePathway = TRUE)

# Collapse smgbp genes into pathway-level mutation rows.
pathways_smgbp <- oncoplot(maf= maf_tumor, pathways = "smgbp", gene_mar =4, fontSize=0.6, topPathways= 10, top = 84,
        sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = TRUE,
        showTumorSampleBarcodes = TRUE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        annotationFontSize = 1.1,
        legendFontSize = 0.9,
        titleFontSize = 1.3,
        SampleNamefontSize = 0.75, collapsePathway = TRUE)

# Collapse Robert categories into pathway-level mutation rows.
pathways_rob <- oncoplot(maf= maf_tumor, pathways = Robert_pathways, gene_mar =4, fontSize=0.5,topPathways= 11, 
        top = 84, sampleOrder = ordered_samples,
        clinicalFeatures = c("HPV_TP53_mutational_group"),
        sortByAnnotation = TRUE,
        showTumorSampleBarcodes = TRUE,
        removeNonMutated = FALSE,
        annotationColor = oncoplot_ann_colors,
        annotationFontSize = 1.1,
        legendFontSize = 0.9,
        titleFontSize = 1.3,
        SampleNamefontSize = 0.75, collapsePathway = TRUE)

pathways_sigpw
#pathways_smgbp
#pathways_rob

```


# 3. Pathway Boxplot Workflow

The sections below merge the pathway boxplot workflow from `9boxplot_pathways_sigpw.Rmd` into this notebook. The code still uses the internal Sanchez-Vega signalling pathway database used by `maftools` when `pathways = "sigpw"`.

# 3.1 Groups and colors

These are the three groups used in the comparison. Keeping the order
fixed makes the tables and plots easier to read.

```{r groups-and-colors}
# Define fixed colors for the three HPV/TP53 groups.
group_colors <- c(
  "HPV-A" = "#377EB8",
  "HPV-I/TP53mut" = "#E69F00",
  "HPV-I/TP53wt" = "#009E73"
)

# Keep only the sample barcode and group columns used in pathway comparisons.
sample_groups <- clinical_for_maf %>%
  select(Tumor_Sample_Barcode, HPV_TP53_mutational_group) %>%
  mutate(
    HPV_TP53_mutational_group = factor(
      HPV_TP53_mutational_group,
      levels = group_order #from 4maf_obj
    )
  )
```

# 3.2 Load maftools sigpw pathways

`maftools::oncoplot(pathways = "sigpw")` does not require a
user-provided pathway table. In this installed version of `maftools`,
`sigpw` is not exposed as a normal `data("sigpw")` object; it is stored
as an internal TSV file in the package `extdata` directory.

The chunk below reads the same internal TSV file used by `maftools`.
The final table is converted to the same two-column format expected by the old
workflow: `gene` and
`general_category`.

```{r load-sigpw-pathways}
# Load internal maftools signalling pathways database (sigpw)
sigpw_file <- system.file(
  "extdata",
  "oncogenic_sig_patwhays.tsv",
  package = "maftools"
)

sigpw_pathways <- read.delim(sigpw_file) %>%
  dplyr::transmute(
    gene = Gene,
    general_category = Pathway
  )

# Count how many database genes are assigned to each pathway.
sigpw_pathways %>%
  count(general_category, name = "n_genes") %>%
  arrange(desc(n_genes))
```

# 3.3 Mutation counts per sample and pathway

Each mutation row in `maf_tumor@data` is joined to its pathway using
`Hugo_Symbol`. `sigpw` can contain a gene in more than one pathway, so
each gene-pathway row is kept and contributes to the corresponding
pathway count.

Samples with no mutation in a pathway are kept with `Num_Mutations = 0`.
This is important because every pathway is compared using the same
sample set.

`general_category` stores the `sigpw` pathway name.

```{r pathway-counts}
# Match tumor mutation records to sigpw pathway annotations.
mutated_sample_pathways <- maf_tumor@data %>%
  as_tibble() %>%
  select(Tumor_Sample_Barcode, Hugo_Symbol) %>%
  inner_join(sigpw_pathways, by = c("Hugo_Symbol" = "gene")) %>%
  filter(general_category %in% sigpw_pathways$general_category)

# Build a complete sample-by-pathway grid so zero-count pathways are retained.
sample_pathway_grid <- expand.grid(
  Tumor_Sample_Barcode = sample_groups$Tumor_Sample_Barcode,
  general_category = unique(sigpw_pathways$general_category),
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  left_join(sample_groups, by = "Tumor_Sample_Barcode")

# Count mutation records per sample and pathway.
pathway_counts <- mutated_sample_pathways %>%
  count(Tumor_Sample_Barcode, general_category, name = "Num_Mutations") %>%
  right_join(sample_pathway_grid, by = c("Tumor_Sample_Barcode", "general_category")) %>%
  mutate(
    Num_Mutations = if_else(is.na(Num_Mutations), 0L, Num_Mutations),
    HPV_TP53_mutational_group = factor(
      HPV_TP53_mutational_group,
      levels = group_order
    )
  ) %>%
  select(
    Tumor_Sample_Barcode,
    HPV_TP53_mutational_group,
    general_category,
    Num_Mutations
  )

# Summarize total mutation counts by pathway.
pathway_counts %>%
  count(general_category, wt = Num_Mutations, name = "total_mutations") %>%
  arrange(desc(total_mutations))
```

# 3.4 Boxplots and Kruskal-Wallis tests

For each pathway, the boxplot shows the number of mutations per sample
in each HPV/TP53 group. The Kruskal-Wallis test compares the three
groups without assuming normal distributions.

```{r pathway-boxplots, fig.width=12, fig.height=8, dpi=300}
# Select pathways with at least one observed mutation.
plot_pathways <- pathway_counts %>%
  group_by(general_category) %>%
  summarise(total_mutations = sum(Num_Mutations), .groups = "drop") %>%
  filter(total_mutations > 0) %>%
  arrange(desc(total_mutations)) %>%
  pull(general_category)

plots <- list()
stats <- list()

# Generate one boxplot and Kruskal-Wallis result per pathway.
for (pathway_name in plot_pathways) {
  data_summary <- pathway_counts %>%
    filter(general_category == pathway_name)

  kw_result <- kruskal.test(
    Num_Mutations ~ HPV_TP53_mutational_group,
    data = data_summary
  )

  stats[[pathway_name]] <- data.frame(
    general_category = pathway_name,
    p_value = kw_result$p.value
  )

  y_max <- max(data_summary$Num_Mutations)

  plots[[pathway_name]] <- ggplot(
    data_summary,
    aes(
      x = HPV_TP53_mutational_group,
      y = Num_Mutations,
      fill = HPV_TP53_mutational_group
    )
  ) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.55) +
    geom_jitter(width = 0.12, size = 1.4, alpha = 0.8, shape = 21, color = "grey20") +
    scale_fill_manual(values = group_colors, drop = FALSE) +
    scale_y_continuous(breaks = seq(0, y_max + 1, by = 1)) +
    labs(
      title = paste("Pathway:", pathway_name),
      x = NULL,
      y = "Number of mutations"
    ) +
    annotate(
      "text",
      x = 2,
      y = y_max + 0.8,
      label = paste("KW p =", signif(kw_result$p.value, 3)),
      size = 3
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold", size = 9)
    )
}

# Combine pathway-level p-values and apply BH correction.
stats_table <- bind_rows(stats) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adj)

stats_table

grid.arrange(grobs = plots, ncol = 3)
```

Two pathways show significant differences between groups after BH
adjustment: - TP53 - NRF2

# 3.5 Supplementary pairwise pathway comparisons

The supplementary analyses below reuse the existing `pathway_counts`
table and the same pathway-level variable, `Num_Mutations`. Because
these are numeric pathway-level counts shown as boxplots, each two-group
comparison uses a Wilcoxon rank-sum test, followed by Benjamini-Hochberg
correction across pathways.

```{r supplementary-pairwise-helper}
# Helper used only for the supplementary pairwise pathway boxplots below.
make_pairwise_pathway_boxplots <- function(
  pathway_data,
  comparison_col,
  comparison_levels,
  comparison_colors,
  comparison_label
) {
  # Restrict pathway counts to the requested comparison groups.
  pairwise_data <- pathway_data %>%
    filter(.data[[comparison_col]] %in% comparison_levels) %>%
    mutate(
      comparison_group = factor(
        .data[[comparison_col]],
        levels = comparison_levels
      )
    )

  # Keep only pathways with mutations in the selected comparison.
  pairwise_pathways <- pairwise_data %>%
    group_by(general_category) %>%
    summarise(total_mutations = sum(Num_Mutations), .groups = "drop") %>%
    filter(total_mutations > 0) %>%
    arrange(desc(total_mutations)) %>%
    pull(general_category)

  pairwise_plots <- list()
  pairwise_stats <- list()

  # Create pathway-specific Wilcoxon tests and boxplots.
  for (pathway_name in pairwise_pathways) {
    data_summary <- pairwise_data %>%
      filter(general_category == pathway_name)

    group_counts <- table(data_summary$comparison_group)

    if (length(group_counts) == 2 && all(group_counts > 0)) {
      wilcox_result <- wilcox.test(
        Num_Mutations ~ comparison_group,
        data = data_summary,
        exact = FALSE
      )
      p_value <- wilcox_result$p.value
    } else {
      p_value <- NA_real_
    }

    pairwise_stats[[pathway_name]] <- data.frame(
      general_category = pathway_name,
      comparison = comparison_label,
      group_1 = comparison_levels[1],
      group_2 = comparison_levels[2],
      n_group_1 = unname(group_counts[comparison_levels[1]]),
      n_group_2 = unname(group_counts[comparison_levels[2]]),
      median_group_1 = median(
        data_summary$Num_Mutations[data_summary$comparison_group == comparison_levels[1]],
        na.rm = TRUE
      ),
      median_group_2 = median(
        data_summary$Num_Mutations[data_summary$comparison_group == comparison_levels[2]],
        na.rm = TRUE
      ),
      p_value = p_value
    )

    y_max <- max(data_summary$Num_Mutations)

    pairwise_plots[[pathway_name]] <- ggplot(
      data_summary,
      aes(
        x = comparison_group,
        y = Num_Mutations,
        fill = comparison_group
      )
    ) +
      geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.55) +
      geom_jitter(width = 0.12, size = 1.4, alpha = 0.8, shape = 21, color = "grey20") +
      scale_fill_manual(values = comparison_colors, drop = FALSE) +
      scale_y_continuous(breaks = seq(0, y_max + 1, by = 1)) +
      labs(
        title = paste("Pathway:", pathway_name),
        x = NULL,
        y = "Number of mutations"
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max + 0.8,
        label = paste("Wilcoxon p =", signif(p_value, 3)),
        size = 3
      ) +
      theme_classic(base_size = 10) +
      theme(
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", size = 9)
      )
  }

  # Combine pairwise statistics and apply BH correction.
  pairwise_stats_table <- bind_rows(pairwise_stats) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
    arrange(p_adj)

  list(
    plots = pairwise_plots,
    stats = pairwise_stats_table
  )
}
```

# 3.6 Supplementary: HPVi vs HPVa pathway boxplots

This comparison collapses the two HPV-independent TP53 subgroups into
one `HPVi` group and compares it against the HPV-associated `HPVa`
group.

```{r supplementary-hpvi-vs-hpva, fig.width=12, fig.height=8, dpi=300}
# Collapse HPV-independent TP53 subgroups into one HPV-I comparison group.
pathway_counts_HPVi_vs_HPVa <- pathway_counts %>%
  mutate(
    HPV_status_pairwise = case_when(
      HPV_TP53_mutational_group == "HPV-A" ~ "HPV-A",
      HPV_TP53_mutational_group %in% c(
        "HPV-I/TP53mut", "HPV-I/TP53wt") ~ "HPV-I",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(HPV_status_pairwise))

# Build HPV-I versus HPV-A pathway plots and statistics.
pathway_pairwise_HPVi_vs_HPVa_result <- make_pairwise_pathway_boxplots(
  pathway_data = pathway_counts_HPVi_vs_HPVa,
  comparison_col = "HPV_status_pairwise",
  comparison_levels = c("HPV-I", "HPV-A"),
  comparison_colors = c(
    "HPV-I" = "#E69F00",
    "HPV-A" = "#377EB8"
  ),
  comparison_label = "HPV-I vs HPV-A"
)

pathway_boxplots_HPVi_vs_HPVa <- pathway_pairwise_HPVi_vs_HPVa_result$plots
pathway_pairwise_HPVi_vs_HPVa <- pathway_pairwise_HPVi_vs_HPVa_result$stats

pathway_pairwise_HPVi_vs_HPVa

grid.arrange(grobs = pathway_boxplots_HPVi_vs_HPVa, ncol = 3)
```

# 3.7 Supplementary: HPVi TP53 wild-type vs HPVi TP53-mutant pathway boxplots

This comparison keeps only HPV-independent samples and compares TP53
wild-type against TP53-mutant tumors.

```{r supplementary-hpvi-tp53wt-vs-hpvi-tp53mut, fig.width=12, fig.height=8, dpi=300}
# Keep only HPV-independent samples and define TP53-status comparison groups.
pathway_counts_HPVi_TP53wt_vs_HPVi_TP53mut <- pathway_counts %>%
  mutate(
    HPVi_TP53_pairwise = case_when(
      HPV_TP53_mutational_group == "HPV-I/TP53wt" ~ "HPV-I/TP53wt",
      HPV_TP53_mutational_group == "HPV-I/TP53mut" ~ "HPV-I/TP53mut",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(HPVi_TP53_pairwise))

# Build HPV-I TP53-wild-type versus TP53-mutant pathway plots and statistics.
pathway_pairwise_HPVi_TP53wt_vs_HPVi_TP53mut_result <- make_pairwise_pathway_boxplots(
  pathway_data = pathway_counts_HPVi_TP53wt_vs_HPVi_TP53mut,
  comparison_col = "HPVi_TP53_pairwise",
  comparison_levels = c("HPV-I/TP53wt", "HPV-I/TP53mut"),
  comparison_colors = c(
    "HPV-I/TP53wt" = "#009E73",
    "HPV-I/TP53mut" = "#E69F00"
  ),
  comparison_label = "HPV-I/TP53wt vs HPV-I/TP53mut"
)

pathway_boxplots_HPVi_TP53wt_vs_HPVi_TP53mut <- pathway_pairwise_HPVi_TP53wt_vs_HPVi_TP53mut_result$plots
pathway_pairwise_HPVi_TP53wt_vs_HPVi_TP53mut <- pathway_pairwise_HPVi_TP53wt_vs_HPVi_TP53mut_result$stats

pathway_pairwise_HPVi_TP53wt_vs_HPVi_TP53mut

grid.arrange(grobs = pathway_boxplots_HPVi_TP53wt_vs_HPVi_TP53mut, ncol = 3)
```

