---
title: "Lollipop Plots and Pathway Summaries"
author: "Elena Diaz"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

# Custom Functions Defined in This Notebook

- `top_mutated_genes()`: returns the top mutated genes from a MAF object, using
  `maftools::getGeneSummary()`.
- `save_lollipop_plot()`: saves a lollipop plot to PNG and skips/removes the
  file if plotting fails.
- `show_lollipop_plot()`: displays a lollipop plot inline while safely skipping
  genes that cannot be plotted.

# Workflow Documentation

## Purpose

This notebook creates protein-domain lollipop plots for recurrently mutated genes in the PSCC tumor MAF object and compares HPV-independent with HPV-associated tumors.

## Required Objects in the R Environment

This notebook expects the following objects:

-   `maf_tumor`: tumor-only MAF object.
-   `maf_hpvi`: HPV-independent tumor MAF object.
-   `maf_hpva`: HPV-associated tumor MAF object.

The MAF objects must contain an amino-acid change column used by `AACol` in the lollipop plots. The preferred column is `hgvs_p`; if needed, the setup chunk also checks `HGVSp_Short`, `Protein_Change`, and `AAChange`.

The current clinical group annotation is `HPV_TP53_mutational_group`, with these levels:

-   `HPV-A`
-   `HPV-I/TP53mut`
-   `HPV-I/TP53wt`


## Main Outputs

PNG files are written to:

`C:/Users/ediaz/pscc_project/outputs/lollipops`

# 1. Setup

```{r lollipop-setup}
# Load plotting library and prepare the lollipop output directory.
library(maftools)

lollipop_dir <- "C:/Users/ediaz/pscc_project/outputs/lollipops"

if (!dir.exists(lollipop_dir)) {
  dir.create(lollipop_dir, recursive = TRUE)
}

required_maf_objects <- c("maf_tumor", "maf_hpvi", "maf_hpva")
missing_maf_objects <- setdiff(required_maf_objects, ls(envir = .GlobalEnv))

# Stop early if any required MAF object is missing from the session.
if (length(missing_maf_objects) > 0) {
  stop(
    "Missing required MAF object(s): ",
    paste(missing_maf_objects, collapse = ", ")
  )
}

preferred_aa_columns <- c("hgvs_p", "HGVSp_Short", "Protein_Change", "AAChange")
aa_column <- preferred_aa_columns[preferred_aa_columns %in% names(maf_tumor@data)][1]

# Choose the amino-acid column used by all lollipop plots.
if (is.na(aa_column)) {
  stop(
    "No amino-acid change column found in maf_tumor. Expected one of: ",
    paste(preferred_aa_columns, collapse = ", ")
  )
}

missing_aa_columns <- required_maf_objects[
  !vapply(
    required_maf_objects,
    function(object_name) aa_column %in% names(get(object_name)@data),
    logical(1)
  )
]

# Confirm that the selected amino-acid column exists in every MAF object.
if (length(missing_aa_columns) > 0) {
  stop(
    "The amino-acid column `", aa_column, "` is missing from: ",
    paste(missing_aa_columns, collapse = ", ")
  )
}

message("Using amino-acid column for lollipop plots: ", aa_column)

# Optional: add genes manually here.
manual_genes_tumor <- c()       # example: c("TP53", "PIK3CA")
manual_genes_hpv <- c()         # example: c("TP53", "NOTCH1")
```

```{r helper-functions}
# Return the most frequently mutated genes for a given MAF object.
top_mutated_genes <- function(maf_object, n = 15) {
  gene_summary <- getGeneSummary(maf_object)

  if (nrow(gene_summary) == 0) {
    return(character())
  }

  head(gene_summary$Hugo_Symbol, n)
}

# Save a plot expression to a PNG file and remove incomplete output on error.
save_lollipop_plot <- function(filename, plot_expression, width, height) {
  plot_ok <- FALSE

  png(
    filename = filename,
    width = width,
    height = height,
    units = "in",
    res = 300
  )

  tryCatch(
    {
      force(plot_expression)
      plot_ok <- TRUE
    },
    error = function(error) {
      message("Skipping plot `", basename(filename), "`: ", error$message)
    },
    finally = {
      dev.off()
    }
  )

  if (!plot_ok && file.exists(filename)) {
    unlink(filename)
  }

  invisible(plot_ok)
}

# Print a plot expression inline while keeping the notebook running on errors.
show_lollipop_plot <- function(plot_expression) {
  tryCatch(
    {
      force(plot_expression)
      invisible(TRUE)
    },
    error = function(error) {
      message("Skipping inline plot: ", error$message)
      invisible(FALSE)
    }
  )
}
```

# 2. Tumor Lollipop Plots

Single lollipop plots are created for the top 15 mutated genes in `maf_tumor`. Additional genes can be added in `manual_genes_tumor`.

```{r lollipop-top15-tumor, fig.width=12, fig.height=6, dpi=300}
# Build the tumor gene list from top mutated genes plus any manual additions.
top15_tumor_genes <- top_mutated_genes(maf_tumor, n = 15)
genes_tumor <- unique(c(top15_tumor_genes, manual_genes_tumor))

# Save and display one lollipop plot per selected tumor gene.
for (gene_name in genes_tumor) {
  message("Plotting tumor lollipop for: ", gene_name)

  save_lollipop_plot(
    filename = file.path(lollipop_dir, paste0(gene_name, "_tumor.png")),
    width = 12,
    height = 6,
    plot_expression = lollipopPlot(
      maf = maf_tumor,
      gene = gene_name,
      AACol = aa_column,
      showMutationRate = TRUE, 
      labelPos = "all"
    )
  )

  show_lollipop_plot(
    lollipopPlot(
      maf = maf_tumor,
      gene = gene_name,
      AACol = aa_column,
      showMutationRate = TRUE, 
      labelPos = "all"
    )
  )
}
```

# 3. HPV-Independent vs HPV-Associated Lollipop Plots

Comparative plots are created for the top 15 mutated genes in `maf_hpvi`, then compared with `maf_hpva`. Additional genes can be added in `manual_genes_hpv`.

```{r lollipop-top15-hpv-comparison, fig.width=15, fig.height=8, dpi=300}
# Build the HPV comparison gene list from HPV-independent tumors plus manual additions.
top15_hpv_genes <- top_mutated_genes(maf_hpvi, n = 15)
genes_hpv <- unique(c(top15_hpv_genes, manual_genes_hpv))

# Save and display comparative lollipop plots for HPV-I versus HPV-A tumors.
for (gene_name in genes_hpv) {
  message("Plotting HPV comparison lollipop for: ", gene_name)

  save_lollipop_plot(
    filename = file.path(lollipop_dir, paste0(gene_name, "_hpvi_vs_hpva.png")),
    width = 15,
    height = 8,
    plot_expression = lollipopPlot2(
      m1 = maf_hpvi,
      m2 = maf_hpva,
      gene = gene_name,
      AACol1 = aa_column,
      AACol2 = aa_column,
      m1_name = "PSCC HPV-independent",
      m2_name = "PSCC HPV-associated",
      m1_label = "none",
      m2_label = "none",
      verbose = FALSE
    )
  )

  show_lollipop_plot(
    lollipopPlot2(
      m1 = maf_hpvi,
      m2 = maf_hpva,
      gene = gene_name,
      AACol1 = aa_column,
      AACol2 = aa_column,
      m1_name = "PSCC HPV-independent",
      m2_name = "PSCC HPV-associated",
      m1_label = "none",
      m2_label = "none",
      verbose = FALSE
    )
  )
}
```



