---
title: "more_metrics_maf"
author: "Elena Díaz"
date: "2026-05-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Custom functions used in this notebook

No custom functions are defined in this notebook. It uses objects and helper
functions created upstream:

- `maf_tumor`: tumor-only MAF object used for mutation metrics.
- `clinical_for_maf`: tumor clinical annotation table used for group summaries.
- `list_analyzed`: filtered variant table used for gene and HGVS counts.
- `save_flextable_docx()`: exports metric tables as Word documents.

## Workflow overview

This notebook calculates tumor mutation burden, mutation count summaries,
variant recurrence metrics, variant class counts, SNV substitution classes, and
top mutated genes for the tumor MAF object.

#TMB Calculations
Notes:
- Capture Size: defines the genomic length covered by your sequencing panel in megabases (using the KAPA HyperExome capture target size of 43 Mb). 
- Variant Type: TMB calculations typically include only non-synonymous mutations (like missense and nonsense variants), while silent/synonymous mutations are often excluded.
- Median Cutoff: Stratify patient cohorts into "TMB-High" (above the median) and "TMB-Low" (at or below the median) groups for immunotherapy response analysis

```{r}
# Estimate tumor mutation burden using the tumor-only MAF object.
#for tumor
# 1. Load or read your MAF object 
# 2. Estimate TMB (default capture size for exome-seq is 50 Mb)
tmb_table <- tmb(maf = maf_tumor, captureSize = 43, logScale = FALSE)
tmb_table

# Summarize the cohort-level median TMB.
# 3. Calculate the median TMB value
median_tmb <- median(tmb_table$total_per_MB)

print(median_tmb)

```

```{r}
# Join TMB values to clinical annotations for group-level comparisons.
#tmb differences in HPV_tP53_mutational_group
tmb_with_clinical <- tmb_table %>%
  inner_join(clinical_for_maf, by = "Tumor_Sample_Barcode")

# Calculate median TMB and sample counts by HPV/TP53 group.
median_per_group <- tmb_with_clinical %>%
  group_by(HPV_TP53_mutational_group) %>%
  summarise(
    Median_TMB = median(total_perMB, na.rm = TRUE),
    Sample_Count = n()
  )

print(median_per_group)

# Test TMB differences across the three HPV/TP53 groups.
# Test de Kruskal-Wallis para 3 grupos
test_global <- kruskal.test(total_perMB ~ HPV_TP53_mutational_group, data = tmb_with_clinical)

# Mostrar el p-valor global
print(test_global$p.value)

# Test TMB differences between HPV-associated and HPV-independent tumors.
#Comparison per two groups HPV
test_results_hpv <- wilcox.test(total_perMB ~ hpv_status, data = tmb_with_clinical)

print(test_results_hpv$p.value)
```
no group significant

# extract all somatic mutations
```{r}
# Summarize the tumor MAF and extract per-sample mutation counts.
plotmafSummary(maf = maf_tumor) #general
sample_summary <- getSampleSummary(maf_tumor)

# Calculate the mean and observed range of somatic mutations per tumor.
mean_mutations <- mean(sample_summary$total)
range_mutations <- range(sample_summary$total)

cat(paste0("The mean number of somatic mutations detected per tumor was ", 
           round(mean_mutations, 2), 
           " (range, ", range_mutations[1], "-", range_mutations[2], ")\n"))
```

#genes in filtered_variants
```{r}
# Count unique mutated genes among tumor samples in the filtered variant table.
#n of unique genes in all filtered_Variants of tumor
list_analyzed %>%
  filter(tissue_type == "tumor") %>%
  pull(gene) %>%
  unique() %>%
  length()

# Count how many tumor samples contain variants in each gene.
gene_counts <- list_analyzed %>%
  filter(tissue_type == "tumor") %>%
  count(gene, name = "n_samples")

gene_counts %>%
  count(n_samples, name = "n_genes") %>%
  arrange(n_samples)
```

# Count how many times each HGVS coding variant appears across the dataset
```{r}
# Count each coding HGVS variant once per sample.
hgvs_counts <- list_analyzed %>%
  filter(tissue_type == "tumor") %>%
  filter(!is.na(hgvs_c), hgvs_c != "") %>%
  distinct(sample_id, hgvs_c) %>%
  count(hgvs_c, name = "n_samples")

# Classify coding variants as unique or recurrent across samples.
hgvs_counts %>%
  mutate(type = ifelse(n_samples == 1, "Unique", "Recurrent")) %>%
  count(type) %>%
  mutate(percent = round(100 * n / sum(n), 1))
```

#nucleotide changes by ai (maybe simpler possible)
```{r}
# Count maftools variant classifications in the tumor MAF.
# Variant_Classification class counts
variant_classification_counts <- maf_tumor@data %>%
  count(Variant_Classification, sort = TRUE)
variant_classification_counts

# Count broad variant types in the tumor MAF.
# Variant_Type class counts
variant_type_counts <- maf_tumor@data %>%
  count(Variant_Type, sort = TRUE)
variant_type_counts

# Export variant class and type count tables.
save_flextable_docx(variant_classification_counts, "variant_classification_counts")
save_flextable_docx(variant_type_counts, "variant_type_counts")
```

```{r}
# Create normalized SNV substitution classes from reference and alternate alleles.
# SNV class counts, IT DOES NOT Exit. Create from ref and alt. Only for SNPs
maf_tumor@data <- maf_tumor@data %>%
  mutate(
    SNV_Class = paste0(ref, ">", alt),

    # Normalize complementary substitutions to pyrimidine context
    SNV_Class_norm = recode(
      SNV_Class,
      "G>A" = "C>T",
      "A>G" = "T>C",
      "G>C" = "C>G",
      "A>C" = "T>G",
      "G>T" = "C>A",
      "A>T" = "T>A",
      .default = SNV_Class
    )
  )

# Count normalized SNP substitution classes and calculate percentages.
# Count normalized SNP classes
snv_class_counts <- maf_tumor@data %>%
  count(SNV_Class_norm, sort = TRUE) %>%
  filter(nchar(SNV_Class_norm) == 3) %>%
  mutate(
    percentage = round(100 * n / sum(n), 2)
  )

# Export normalized SNV substitution counts.
save_flextable_docx(snv_class_counts, "snv_class_counts")
```

Study the most mutated genes (by total variants, and total samples affecting)
- Hugo_Symbol: Name of gene
- MutatedSamples: Number of patients/samples in whcih gene is mutated.
- total: Total number of mutational events in that gene (equal/higher to mutated samples)
```{r}
# Extract the most frequently mutated genes by total mutation events.
#table for top 10 mutated genes. Cols: genes, total variants, unique per sample
top_genes_table <- getGeneSummary(maf_tumor) %>%
  select(Hugo_Symbol, MutatedSamples, total) %>%
  arrange(desc(total)) %>%
  head(20) 

top_genes_table

# Export the top mutated genes table.
save_flextable_docx(top_genes_table, "top_mutated_genes")
```
