---
title: "Oncoplot Generation"
author: "Elena Díaz"
date: "2026-05-11"
output: html_document
---

## Custom functions used in this notebook

No custom functions are defined in this notebook. It uses helper objects and
functions created upstream:

- `top_genes()`: selects the most frequently mutated genes from a MAF object.
- `get_ordered_genes()`: returns genes in the order used for grouped tumor
  oncoplots.
- `save_flextable_docx()`: exports comparison tables as Word documents.

## Workflow overview

This notebook generates summary plots, oncoplots, group comparisons, and final
MAF exports from the MAF objects prepared in `4maf_obj_creation.Rmd`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r tumor-premalignant, fig.width=17, fig.height=10, dpi=300, fig.retina=2}
# Summarize mutation burden and variant classes for tumor and premalignant samples.
summary_tumor_premalignant <- plotmafSummary(
  maf = maf_tumor_premalignant,
  rmOutlier = TRUE,
  addStat = "median",
  dashboard = TRUE,
  titvRaw = FALSE
)
summary_tumor_premalignant

# Generate the combined tumor/premalignant oncoplot with clinical annotations.
oncoplot_tumor_premalignant <- oncoplot(
  maf = maf_tumor_premalignant,
  top = 30,
  clinicalFeatures = c(
    "HPV_TP53_mutational_group",
    "tissue_type",
    "tp53_status",
    "p53_ihc_status",
    "cyclind1_ihc_perc_grp",
    "matched"
  ),
  sortByAnnotation = TRUE,
  removeNonMutated = FALSE,
  showTumorSampleBarcodes = TRUE,
  annotationColor = oncoplot_ann_colors,
  annotationFontSize = 1.15,
  legendFontSize = 0.9,
  titleFontSize = 1.3,
  fontSize = 0.6,
  SampleNamefontSize = 0.7,
  barcode_mar = 5,
  gene_mar = 8
)
oncoplot_tumor_premalignant
```

# 6. Tumor-Only Analysis

```{r tumor-oncoplot, fig.width=18, fig.height=12, dpi=300, fig.retina=2}
# Summarize mutation patterns for tumor samples only.
plotmafSummary(
  maf = maf_tumor,
  rmOutlier = TRUE,
  addStat = "median",
  dashboard = TRUE,
  titvRaw = FALSE
)

# Generate the tumor-only oncoplot
oncoplot(
  maf = maf_tumor,
  top = 84,
  clinicalFeatures = c(
    "HPV_TP53_mutational_group",
    "tp53_status",
    "p53_ihc_status",
    "cyclind1_ihc_perc_grp",
    "matched"
  ),
  sortByAnnotation = TRUE,
  removeNonMutated = FALSE,
  showTumorSampleBarcodes = TRUE,
  annotationColor = oncoplot_ann_colors,
  annotationFontSize = 1.2,
  legendFontSize = 0.9,
  titleFontSize = 1.3,
  fontSize = 0.45,
  SampleNamefontSize = 0.75,
  barcode_mar = 5,
  gene_mar = 9
)
```

# 7. HPV-a vs HPV-i

```{r hpv-subset-oncoplots, fig.width=12, fig.height=8, dpi=300, fig.retina=2}
# Plot the HPV-associated tumor subset.
plotmafSummary(maf = maf_hpva)
oncoplot(
  maf = maf_hpva,
  top = 50,
  clinicalFeatures = c(
    "tp53_status",
    "p53_ihc_status",
    "cyclind1_ihc_perc_grp",
    "matched"
  ),
  sortByAnnotation = TRUE,
  removeNonMutated = FALSE,
  showTumorSampleBarcodes = TRUE,
  annotationColor = oncoplot_ann_colors,
  annotationFontSize = 1.5,
  legendFontSize = 1,
  titleFontSize = 1.5,
  fontSize = 0.7,
  SampleNamefontSize = 1,
  barcode_mar = 8,
  gene_mar = 8
)

# Plot the HPV-independent tumor subset.
plotmafSummary(maf = maf_hpvi)
oncoplot(
  maf = maf_hpvi,
  top = 50,
  clinicalFeatures = c(
    "tp53_status",
    "p53_ihc_status",
    "cyclind1_ihc_perc_grp",
    "matched"
  ),
  sortByAnnotation = TRUE,
  removeNonMutated = FALSE,
  showTumorSampleBarcodes = TRUE,
  annotationColor = oncoplot_ann_colors,
  annotationFontSize = 1.5,
  legendFontSize = 1,
  titleFontSize = 1.5,
  fontSize = 0.7,
  SampleNamefontSize = 1,
  barcode_mar = 5,
  gene_mar = 8
)
```

# 8. HPV-Associated vs HPV-Independent Comparison
Table of comparsion

```{r hpva-hpvi-comparison}
# Compare gene mutation frequencies between HPV-associated and HPV-independent tumors.
hpva_vs_hpvi <- mafCompare(
  m2 = maf_hpva,
  m1 = maf_hpvi,
  m2Name = "HPV-associated",
  m1Name = "HPV-independent",
  minMut = 3
)

# Prepare an interactive comparison table and export it for reporting.
hpva_vs_hpvi_results <- as.data.frame(hpva_vs_hpvi$results)
colnames(hpva_vs_hpvi_results) <- make.names(colnames(hpva_vs_hpvi_results), unique = TRUE)

hpva_vs_hpvi_results_table <- hpva_vs_hpvi_results %>%
  datatable(filter = list(position = "top", clear = FALSE)) %>%
  formatStyle("pval", color = styleInterval(c(0.05), c("darkgreen", "black")))
hpva_vs_hpvi_results_table

save_flextable_docx(hpva_vs_hpvi_results, "hpv_mafcompare_results")
```

```{r hpva-hpvi-comparison-plots, fig.width=10, fig.height=7, dpi=300, fig.retina=2}
# Select genes for visual comparison using top genes plus near-significant results.
hpva_vs_hpvi_genes <- unique(c(
  top_genes(maf_hpva, 10),
  top_genes(maf_hpvi, 10),
  hpva_vs_hpvi_results$Hugo_Symbol[hpva_vs_hpvi_results$pval < 0.08]
)) 

# Compare mutation patterns between HPV-associated and HPV-independent tumors.
hpva_vs_hpvi_coOncoplot <- coOncoplot(
  m1 = maf_hpva,
  m2 = maf_hpvi,
  m1Name = "HPV-associated",
  m2Name = "HPV-independent",
  genes = hpva_vs_hpvi_genes,
  removeNonMutated = FALSE,
      legendFontSize = 1,
      annotationFontSize = 1, 
      gene_mar=1
)
hpva_vs_hpvi_coOncoplot

# Compare mutation frequencies between HPV-associated and HPV-independent tumors.
hpva_vs_hpvi_coBarplot <- coBarplot(
  m1 = maf_hpva,
  m2 = maf_hpvi,
  genes = hpva_vs_hpvi_genes,
  m1Name = "HPV-associated",
  m2Name = "HPV-independent",
      titleSize = 0.8,
      geneSize = 0.6,
      showPct = TRUE,
      pctSize = 0.7,
      axisSize = 0.7,
      legendTxtSize = 0.8
)
hpva_vs_hpvi_coBarplot
```

# 10. HPV-Independent TP53-Wild-Type vs TP53-Mutated Comparison

```{r hpvi-tp53-comparison}
# Compare HPV-independent tumors by sequencing-derived TP53 status.
hpvi_tp53wt_vs_tp53mut <- mafCompare(
  m2 = maf_hpvi_tp53wt,
  m1 = maf_hpvi_tp53mut,
  m2Name = "HPVI TP53-wild-type",
  m1Name = "HPVI TP53-mutated",
  minMut = 3
)

# Prepare an interactive comparison table and export it for reporting.
hpvi_tp53_results <- as.data.frame(hpvi_tp53wt_vs_tp53mut$results)
colnames(hpvi_tp53_results) <- make.names(colnames(hpvi_tp53_results), unique = TRUE)

hpvi_tp53_results_table <- hpvi_tp53_results %>%
  datatable(filter = list(position = "top", clear = FALSE)) %>%
  formatStyle("pval", color = styleInterval(c(0.05), c("darkgreen", "black")))
hpvi_tp53_results_table

save_flextable_docx(hpvi_tp53_results, "hpvi_TPstatus_mafcompare_result")

```


```{r hpvi-tp53-comparison-plots, fig.width=10, fig.height=7, dpi=300, fig.retina=2}
# Select genes for visual comparison using top genes plus significant results.
hpvi_tp53_genes <- unique(c(
  top_genes(maf_hpvi_tp53wt, 10),
  top_genes(maf_hpvi_tp53mut, 10),
  hpvi_tp53_results$Hugo_Symbol[hpvi_tp53_results$pval < 0.05]
))

# Compare mutation patterns between HPV-I/TP53-wt and HPV-I/TP53-mut tumors.
hpvi_tp53_coOncoplot <-  coOncoplot(
      m1 = maf_hpvi_tp53wt,
      m2 = maf_hpvi_tp53mut,
      m1Name = "HPV-I/TP53wt",
      m2Name = "HPV-I/TP53mut",
      genes = hpvi_tp53_genes,
      removeNonMutated = TRUE,
      legendFontSize = 0.8,
      annotationFontSize = 0.8,
      gene_mar=1
)
hpvi_tp53_coOncoplot

# Compare mutation frequencies between HPV-I/TP53-wt and HPV-I/TP53-mut tumors.
hpvi_tp53_coBarplot <- coBarplot(
      m1 = maf_hpvi_tp53wt,
      m2 = maf_hpvi_tp53mut,
      genes = hpvi_tp53_genes,
      m1Name = "HPV-I/TP53wt",
      m2Name = "HPV-I/TP53mut",
      titleSize = 0.8,
      geneSize = 0.6,
      showPct = TRUE,
      pctSize = 0.7,
      axisSize = 0.7,
      legendTxtSize = 0.8
)

hpvi_tp53_coBarplot
```

#General Oncoplot for tumor samples 

```{r, fig.width=16, fig.height=10, dpi=300, fig.retina=2}
# Use the full tumor gene set for the grouped general oncoplot.
n_top_genes <- 84

ordered_genes <- get_ordered_genes(n_top_genes)

# Generate the grouped tumor oncoplot using the predefined sample and gene order.
general_oncoplot <- oncoplot(
  maf = maf_three_groups,
  genes = ordered_genes,
  keepGeneOrder = TRUE,
  sampleOrder = ordered_samples,
  clinicalFeatures = c(
    "HPV_TP53_mutational_group",
    "tp53_status",
    "p53_ihc_status",
    "cyclind1_ihc_perc_grp"
  ),
  showTumorSampleBarcodes = TRUE,
  annotationColor = oncoplot_ann_colors,
  barcode_mar = 5,
  gene_mar = 4
)
general_oncoplot
```

Same for top 30 genes
```{r, fig.width=16, fig.height=10, dpi=300, fig.retina=2}
# Repeat the grouped tumor oncoplot for the top 30 ordered genes.
n_top_genes <- 30

ordered_genes <- get_ordered_genes(n_top_genes)

plotmafSummary(maf = maf_three_groups)

general_oncoplot <- oncoplot(
  maf = maf_three_groups,
  genes = ordered_genes,
  keepGeneOrder = TRUE,
  sampleOrder = ordered_samples,
  clinicalFeatures = c(
    "HPV_TP53_mutational_group",
    "tp53_status",
    "p53_ihc_status"
  ),
  showTumorSampleBarcodes = FALSE,
  annotationColor = oncoplot_ann_colors,
  barcode_mar = 2,
  gene_mar = 8
)
general_oncoplot

```

# Export MAF object
```{r export-maf-tumor}
# Create the export directory and save the tumor MAF object plus source tables.
maf_tumor_export_dir <- "maf_tumor_export"
dir.create(maf_tumor_export_dir, showWarnings = FALSE, recursive = TRUE)

save(
  maf_tumor,
  file = file.path(maf_tumor_export_dir, "maf_tumor_full_object.RData")
)
data.table::fwrite(
  maf_tumor@data,
  file = file.path(maf_tumor_export_dir, "maf_tumor_mutations.tsv"),
  sep = "\t"
)
data.table::fwrite(
  maf_tumor@clinical.data,
  file = file.path(maf_tumor_export_dir, "maf_tumor_clinical_data.tsv"),
  sep = "\t"
)

cat("maf_tumor exported to:", normalizePath(maf_tumor_export_dir), "\n")
```
