---
title: "Untitled"
author: "Núria Peñuelas Peñarroya & Elena Diaz"
date: "2026-04-23"
output: html_document
---

## Script documentation

- Custom functions defined in this file: none.
- External helper used: `save_flextable_docx()` saves the formatted Cox regression tables to Word documents.
- Main workflow: load TP53/sample metadata, retrieve clinical REDCap data, prepare 60-month survival variables, define HPV/TP53 groups, plot Kaplan-Meier curves, and run Cox regression models.


```{r}
# Load TP53/sample metadata generated by the filtering pipeline.
tp53 <-read.csv("C:/Users/ediaz/pscc_project/downloads/20260527_Elena_nuevapipeline_nuevosfiltros_tp53status.csv")
```

# Install the packages if not already installed

```{r packages}
# Load packages for REDCap export, data wrangling, survival models, and tables.
library(redcapAPI)
library(gtsummary)
library(magrittr)
library(dplyr)
library(survival)
library(survminer)
library(forcats)
```

#Data loading from REDCAP CADMIAPLUS:

```{r}
# Connect to the CADMIAPLUS REDCap project using local tokens.
source("C:/Users/ediaz/pscc_project/tokens.R")
rcon_cadmiaplus = redcapConnection(url = api_url, token = token_PSCC)
```

```{r}
# Export the clinical variables needed for survival endpoints and covariates.
 clinical <- exportRecordsTyped(rcon = rcon_cadmiaplus,
                                              fields = c('study_id',
                                                        'id_tumoral',
                                                        'diag_date', #instead of surg_date
                                                         'last_follow_up',
                                                         'status_last_followup',
                                                         'recurrence',
                                                         'recurrence_date',
                                                        'p53_ihc_status',
                                                         "age_calc",
                                                        "stage",
                                                        "lymph_node"))

 length(clinical$study_id)
 #155
```

```{r}
# Use the TP53 metadata file as the filtered sample metadata table.
metadata_pscc_filtered<-tp53

# Samples in metadata not present in clinical
missing_in_clinical <- setdiff(metadata_pscc_filtered$sample_id, clinical$id_tumoral)

missing_in_clinical
length(missing_in_clinical)
#0
```

```{r}
# Join filtered sample metadata with clinical REDCap data.
metadata_pscc_clinical <- metadata_pscc_filtered %>%
  left_join(clinical, by = c("Tumor_Sample_Barcode" = "id_tumoral"))
```

```{r}
# Confirm that every sample has matched clinical information.
metadata_pscc_clinical %>%
  filter(is.na(study_id)) %>%
  select(Tumor_Sample_Barcode)
#0
```

```{r}
# Check the number of samples after joining clinical data.
length(metadata_pscc_clinical$Tumor_Sample_Barcode)
#35
#Now there are 33, as Elena told me that in my previous survival there were 2 more cases than she had now(they removed an hypermutated sample... and they should have removed as well another one)
```

#we check if all recurrences have dates
```{r}
# Identify recurrent cases missing a recurrence date.
metadata_pscc_clinical %>%
  filter(recurrence == "Yes" & is.na(recurrence_date))
```

```{r}
# Derive 60-month survival times and event indicators for recurrence and disease-specific death.
metadata_pscc_clinical_60m <- metadata_pscc_clinical %>%
  mutate(
    # --- Convert dates (remove timezone safely) ---
    diag_date    = as.Date(diag_date),
    recurrence_date  = as.Date(recurrence_date),
    last_follow_up   = as.Date(last_follow_up),

    # --- Binary variables ---
    recurrence = ifelse(recurrence == "Yes", 1,
                        ifelse(recurrence == "No", 0, NA)),
    
    #death = ifelse(status_last_followup %in% c("Dead of disease", "Dead of other causes"), 1, 0),
    
    death_disease = ifelse(status_last_followup == "Dead of disease", 1, 0),

    # --- Time calculations (in months) ---
    time_recurrence = as.numeric(difftime(recurrence_date, diag_date, 
                                          units = "days") / (365.25 / 12)),
    
    time_last_followup = as.numeric(difftime(last_follow_up, diag_date, 
                                             units = "days") / (365.25 / 12)),
    
    time_death_surv = time_last_followup,
    
    time_recurrence_surv = ifelse(recurrence ==1 & !is.na(time_recurrence), #added condition extra recurrence =1 to avoid errors of missing recurrence data with recurrence = 0
                                 time_recurrence,
                                 time_last_followup),
 # --- 60-month truncation ---
    time_recurrence_surv_60m = ifelse(time_recurrence_surv <= 60,
                                     time_recurrence_surv, 60),
        time_death_surv_60m = ifelse(time_death_surv <= 60,
                                time_death_surv, 60),
 
        recurrence_60m = ifelse(recurrence == 1 & !is.na(time_recurrence) & 
                        time_recurrence <= 60, 1, 0), #changed
 
        death_disease_60m = ifelse(death_disease == 1 & !is.na(time_death_surv) & 
                           time_death_surv <= 60, 1, 0) #changed
  )   
```

```{r}
# Review survival time distributions and event counts after 60-month truncation.
summary(metadata_pscc_clinical_60m$time_recurrence_surv_60m)
summary(metadata_pscc_clinical_60m$time_death_surv_60m)


table(metadata_pscc_clinical_60m$recurrence)
#7 recurrences in total
table(metadata_pscc_clinical_60m$recurrence_60m)
#7 reucrrences before 5 years
table(metadata_pscc_clinical_60m$death_disease)
#8 disease-specific deaths in total
table(metadata_pscc_clinical_60m$death_disease_60m)
#7 disease-specific death better for analysis...
```

Información clínica importante:
Seguimiento por 60 meses (5 años)
Si el seguimiento es menos a 6 meses y el paciente en el último follow-up está vivo, se descarta
```{r}
# Inspect samples with very short follow-up before applying the exclusion rule.
metadata_pscc_clinical_60m %>%
  filter(time_death_surv_60m < 6) %>%
  select(Tumor_Sample_Barcode, diag_date, last_follow_up, time_last_followup, status_last_followup)
```
#we remove the only patient that has a very short follow-up and has not died
```{r}
# Exclude short follow-up samples only when the patient did not die of disease.
metadata_pscc_clinical_60m <- metadata_pscc_clinical_60m %>%
  filter(time_death_surv_60m >= 6 |
    status_last_followup == "Dead of disease")

length(metadata_pscc_clinical_60m$Tumor_Sample_Barcode)
#[1] 34, we had 34 at the beging but now:
#Now we have 33, the short folowup must have been from a patients that elena also discarded
#We eliminate the patient that is alives but has no follow-up under 6month.We maintain the patient that dies before the 6 month. Then 32.
```

#We create a group based on mutational status of p53 (genomic)

```{r}
# Create genomic HPV/TP53 subgroup labels for survival comparisons.
metadata_pscc_clinical_60m <- metadata_pscc_clinical_60m %>%
  mutate(
    group_genomic = case_when(
      hpv_status == "Associated" ~ "HPV-A",
      hpv_status == "Independent" & TP53_mutational_status == "TP53-wild-type"  ~ "HPV-I/TP53wt",
      hpv_status == "Independent" & TP53_mutational_status == "TP53-mutated" ~ "HPV-I/TP53mut",
      TRUE ~ NA_character_
    )
  )
```

#Disease-specific death and recurrence for HPV/TP53 subgroups (univariate)
```{r}
#dds
# Fit Kaplan-Meier curves for disease-specific survival by genomic subgroup.
fit_km <- survfit(
  Surv(time_death_surv_60m, death_disease_60m) ~ group_genomic, # same as HPV_TP53_mutational_group
  data = metadata_pscc_clinical_60m
)

# Plot disease-specific survival with risk table and log-rank p-value.
plot_dds <- ggsurvplot(
  fit_km,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "DDS",
  legend.title = "HPV/TP53 Subgroup",
  legend.labs = c("HPV-A", "HPV-I/TP53mut", "HPV-I/TP53wt"),
  palette = c("green3", "blue4", "darkred"),  risk.table.height = 0.4,
  # === CONFIGURACIÓN DE TAMAÑOS DE TEXTO (AQUÍ ESTÁ EL CAMBIO) ===
  font.x = c(10, "plain", "black"),          # Título del eje X (Tamaño 10)
  font.y = c(10, "plain", "black"),          # Título del eje Y (Tamaño 10)
  font.tickslab = c(8, "plain", "black"),    # Números de los ejes X e Y (Tamaño 8)
  font.legend = c(8, "plain", "black"),      # Texto de las categorías (Tamaño 8)
  font.title = c(10, "bold", "black"),       # Título del gráfico si tuvieras
  
  # === TAMAÑO DE TEXTO PARA LA TABLA DE RIESGO ===
  tables.theme = theme_survminer(
    font.main = c(9, "bold", "black"),       # Título de la tabla de riesgo
    font.x = c(9, "plain", "black"),         # Título del eje X de la tabla
    font.y = c(9, "plain", "black"),         # Nombres de los grupos en la tabla
    font.tickslab = c(8, "plain", "black")   # Números dentro de la tabla
  )
)
plot_dds
#recurrence
# Fit Kaplan-Meier curves for recurrence-free survival by genomic subgroup.
fit_km <- survfit(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ group_genomic,
  data = metadata_pscc_clinical_60m
)

# Plot recurrence-free survival with risk table and log-rank p-value.
plot_rfs <- ggsurvplot(
  fit_km,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "RFS",
  legend.title = "HPV/TP53 Subgroup",
  legend.labs = c("HPV-A", "HPV-I/TP53mut", "HPV-I/TP53wt"),
  palette = c("green3", "blue4", "darkred"),
  risk.table.height = 0.4,
   # === CONFIGURACIÓN DE TAMAÑOS DE TEXTO (AQUÍ ESTÁ EL CAMBIO) ===
  font.x = c(10, "plain", "black"),          # Título del eje X (Tamaño 10)
  font.y = c(10, "plain", "black"),          # Título del eje Y (Tamaño 10)
  font.tickslab = c(8, "plain", "black"),    # Números de los ejes X e Y (Tamaño 8)
  font.legend = c(8, "plain", "black"),      # Texto de las categorías (Tamaño 8)
  font.title = c(10, "bold", "black"),       # Título del gráfico si tuvieras
  
  # === TAMAÑO DE TEXTO PARA LA TABLA DE RIESGO ===
  tables.theme = theme_survminer(
    font.main = c(9, "bold", "black"),       # Título de la tabla de riesgo
    font.x = c(9, "plain", "black"),         # Título del eje X de la tabla
    font.y = c(9, "plain", "black"),         # Nombres de los grupos en la tabla
    font.tickslab = c(8, "plain", "black")   # Números dentro de la tabla
  )
)

plot_rfs
```

```{r}
#new variable subgroups
# Create IHC-based HPV/p53 subgroup labels.
metadata_pscc_clinical_60m <- metadata_pscc_clinical_60m %>%
  mutate(
    group_ihc = case_when(
      hpv_status == "Associated" ~ "HPV-A",
      hpv_status == "Independent" & p53_ihc_status == "Wt"  ~ "HPV-I/p53-wt",
      hpv_status == "Independent" & p53_ihc_status == "Mutant" ~ "HPV-I/p53-ab",
      TRUE ~ NA_character_
    )
  )
```

#COX REGRESSION
Categorical --> factor variables
```{r}
# Convert categorical variables to factors before Cox modeling.
metadata_pscc_clinical_60m <- metadata_pscc_clinical_60m %>%
  mutate(
    group_genomic = factor(group_genomic),
    hpv_status = factor(hpv_status),
    TP53_mutational_status = factor(TP53_mutational_status),
    stage = factor(stage),
    p53_ihc_status = factor(p53_ihc_status),
    group_ihc = factor(group_ihc)
  )
```

#COX univariate by group_genomic, hpv_status, tp53, stage and lymph node
First we set HPVI/TP53wt as reference group to have comparisons from other groups to it
We only look at the comparison HPVI/TP53wt vs TP53mut
and HPVA vs HPVI/TP53mut. We exclude the third 
SURVIVAL


# gtsummary table — univariable Cox (publication-ready)
summary of all cox regression. 
First we reorder the factors, the reference will be the less agressive
for age <60 vs >=60 years
SURVIVAL
```{r}
# Prepare model-ready covariates and set reference levels for interpretation.
metadata_prepared <- metadata_pscc_clinical_60m %>%
  mutate(
        age_groups = case_when(age_calc < 60  ~ "<60",
                              age_calc >= 60 ~ ">=60"),
        age_groups = fct_relevel(age_groups, "<60"),
#hpv: associated as ref
# HPV_TP53_mutational_group =  "HPV-A"
        TP53_mutational_status = fct_relevel(TP53_mutational_status, "TP53-wild-type"),
        group_ihc = fct_relevel(group_ihc, "HPV-I/p53-wt"),
        stage = case_when(stage =="I"  ~ "Initial (I-II)",
                          stage =="II"  ~ "Initial (I-II)",
                          stage =="III"  ~ "Advanced (III-IV)",
                          stage =="IV"  ~ "Advanced (III-IV)" ),
        stage= fct_relevel(stage, "Initial (I-II)"),
        p53_ihc_status = case_when(p53_ihc_status=="Mutant" ~ "p53-aberrant",
                                 p53_ihc_status=="Wt" ~ "p53-wild-type"),
        p53_ihc_status = fct_relevel(p53_ihc_status, "p53-wild-type")
  )

```


```{r}
# Run univariable Cox regressions for disease-specific survival.
cox_uni_tbl_dds <- tbl_uvregression(
  data = metadata_prepared,
  method = coxph,
  y = Surv(time_death_surv_60m, death_disease_60m),
  exponentiate = TRUE,
  include = c(HPV_TP53_mutational_group, hpv_status, TP53_mutational_status, group_ihc, p53_ihc_status, age_groups, stage),
  pvalue_fun = label_style_pvalue(digits = 2), 
  add_estimate_to_reference_rows = FALSE,
  reference_row = FALSE) %>%
  modify_column_hide(columns = stat_n) %>% #avoid N of variable 
  modify_table_body(
    ~ .x %>%
      relocate(n_obs, .before = estimate)
  ) %>%
  modify_header(
    n_obs ~ "**N**"
  ) %>% #set N of subvariable 
  bold_p(t = 0.05)

# Rename selected table labels for publication-ready output.
cox_uni_tbl_dds <- cox_uni_tbl_dds %>%
  modify_table_body(
    ~ .x %>%
      mutate(
        label = case_when(
          variable == "HPV_TP53_mutational_group" & row_type == "label" ~ "Clinico-pathological subgroups",
          variable == "hpv_status" & row_type == "label" ~ "HPV status",
          variable == "TP53_mutational_status" & row_type == "label" ~ "TP53 mutational status",
          variable == "p53_ihc_status" & row_type == "label" ~ "p53 IHC status",
          variable == "age_groups" & row_type == "label" ~ "Age",
          variable == "stage" & row_type == "label" ~ "Stage",
          TRUE ~ label
        )
      )
  )

cox_uni_tbl_dds
cox_uni_tbl_dds$table_body
#gives problems the function flextable
```
Stage takes into account both metastasis and tumor stage!!
Only 8 events
The only variables that are significant are TP53 mutational status and stage. Then the question we make is:
Are really patients with TP53 mutation have a worse prognosis, or is the bad prognosis just due to advanced stage?

We can check this with a multivariate cox regression including both variables, but we have to be careful with the number of events that we have (7 recurrences and 7 disease-specific deaths) and the number of variables that we include in the model. We can only include 1 variable for every 10 events, so we can only include one variable in the model. 

#multivariate tp53 - stage
```{r}
# Fit a multivariable Cox model with TP53 status and stage for disease-specific survival.
cox_tp53_stage_dds <- coxph(
  Surv(time_death_surv_60m, death_disease_60m) ~ 
    TP53_mutational_status + stage,
  data = metadata_prepared
)

summary(cox_tp53_stage_dds)
```
Both variables tested in multivariate analysis, show that only stage Advanced is significant. 
Conclusions to be said: TP53 mutational status is associated with worse prognosis, but this is due to the fact that TP53 mutations are more frequent in advanced stages(6/10 = 60%) comparing to TP53wt (6/22=27.27%). We cannot conclude that TP53 mutations are an independent prognostic factor in our series, but we can say that they are associated with worse prognosis because they are more frequent in advanced stages.
Why? 
H1. Casuality those patients have advanced - tp53 mut
H2. TP53mut advances tumor progression and patients diagnosed are mostly in Advanced stage.
No se puede saber bien por la baja n (8 eventos)

#UNIVARIATE COX REGRESSION FOR RECURRENCE

```{r}
# Run univariable Cox regressions for recurrence-free survival.
cox_uni_tbl_rec <- tbl_uvregression(
  data = metadata_prepared,
  method = coxph,
  y = Surv(time_recurrence_surv_60m, recurrence_60m),
  exponentiate = TRUE,
  include = c(HPV_TP53_mutational_group, hpv_status, TP53_mutational_status, p53_ihc_status, age_groups, stage),
  pvalue_fun = label_style_pvalue(digits = 2), 
  add_estimate_to_reference_rows = FALSE,
  reference_row = FALSE) %>%
  modify_column_hide(columns = stat_n) %>% #avoid N of variable 
  modify_table_body(
    ~ .x %>%
      relocate(n_obs, .before = estimate)
  ) %>%
  modify_header(
    n_obs ~ "**N**"
  ) %>% #set N of subvariable 
  bold_p(t = 0.05)

# Rename selected recurrence table labels for publication-ready output.
cox_uni_tbl_rec <- cox_uni_tbl_rec %>%
  modify_table_body(
    ~ .x %>%
      mutate(
        label = case_when(
          variable == "HPV_TP53_mutational_group" & row_type == "label" ~ "Clinico-pathological subgroups",
          variable == "hpv_status" & row_type == "label" ~ "HPV status",
          variable == "TP53_mutational_status" & row_type == "label" ~ "TP53 mutational status",
          variable == "p53_ihc_status" & row_type == "label" ~ "p53 IHC status",
          variable == "age_groups" & row_type == "label" ~ "Age",
          variable == "stage" & row_type == "label" ~ "Stage",
          TRUE ~ label
        )
      )
  )

cox_uni_tbl_rec
cox_uni_tbl_rec$table_body

#run save_flextable_function in results gives problems
```
In recurrence only significant at advanced disease. No other variant is significant.
Only 7 events.

#We group
- hpv-a vs hpv-i (no significant)
```{r}
# Compare disease-specific survival by HPV status.
fit_km2 <- survfit(
  Surv(time_death_surv_60m, death_disease_60m) ~ hpv_status, 
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Disease-specific survival probability",
  palette = c("green3", "blue4"),  risk.table.height = 0.4
)

# Compare recurrence-free survival by HPV status.
fit_km2 <- survfit(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ hpv_status,
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Recurrence survival probability",
  palette = c("green3", "blue4"),
  risk.table.height = 0.4
)

# Fit univariable Cox models for HPV status.
cox_group_dds <- coxph(
  Surv(time_death_surv_60m, death_disease_60m) ~ hpv_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_dds)

cox_group_rec <- coxph(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ hpv_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_rec)


```
- p53 ihc (no significant)
```{r}
# Compare disease-specific survival by p53 IHC status.
fit_km2 <- survfit(
  Surv(time_death_surv_60m, death_disease_60m) ~ p53_ihc_status, 
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Disease-specific survival probability",
  palette = c("green3", "blue4"),  risk.table.height = 0.4
)

# Compare recurrence-free survival by p53 IHC status.
fit_km2 <- survfit(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ p53_ihc_status,
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Recurrence survival probability",
  palette = c("green3", "blue4"),
  risk.table.height = 0.4
)

# Fit univariable Cox models for p53 IHC status.
cox_group_dds <- coxph(
  Surv(time_death_surv_60m, death_disease_60m) ~ p53_ihc_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_dds)

cox_group_rec <- coxph(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ p53_ihc_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_rec)

```
- TP53_mutational_status (both significant)

```{r}
# Compare disease-specific survival by TP53 mutational status.
fit_km2 <- survfit(
  Surv(time_death_surv_60m, death_disease_60m) ~ TP53_mutational_status, 
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Disease-specific survival probability",
  palette = c("green3", "blue4"),  risk.table.height = 0.4
)

# Fit disease-specific survival Cox model for TP53 mutational status.
cox_group_dds <- coxph(
  Surv(time_death_surv_60m, death_disease_60m) ~ TP53_mutational_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_dds)

# Compare recurrence-free survival by TP53 mutational status.
fit_km2 <- survfit(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ TP53_mutational_status,
  data = metadata_pscc_clinical_60m
)

ggsurvplot(
  fit_km2,
  data = metadata_pscc_clinical_60m,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Time (months)",
  ylab = "Recurrence survival probability",
  palette = c("green3", "blue4"),
  risk.table.height = 0.4
)

# Fit recurrence-free survival Cox model for TP53 mutational status.
cox_group_rec <- coxph(
  Surv(time_recurrence_surv_60m, recurrence_60m) ~ TP53_mutational_status,
  data = metadata_pscc_clinical_60m
)
summary(cox_group_rec)
```





